% =============================================================================
% Tricopter INDI Flight Controller — Full Technical Report
% =============================================================================
\documentclass[12pt, a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{bm}                % bold math symbols
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, calc}

% Code listing style
\lstdefinestyle{cppstyle}{
    language=C++,
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray}\itshape,
    stringstyle=\color{red},
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    numbersep=5pt,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!5},
    tabsize=4,
    showstringspaces=false,
    captionpos=b,
}
\lstdefinestyle{yamlstyle}{
    basicstyle=\ttfamily\footnotesize,
    commentstyle=\color{gray}\itshape,
    numbers=left,
    numberstyle=\tiny\color{gray},
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!5},
    tabsize=2,
    showstringspaces=false,
    captionpos=b,
}
\lstset{style=cppstyle}

% Shorthand commands
\newcommand{\R}{\mathbb{R}}
\newcommand{\bx}{\bm{x}}
\newcommand{\bu}{\bm{u}}
\newcommand{\bq}{\bm{q}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\btau}{\bm{\tau}}
\newcommand{\bF}{\bm{F}}
\newcommand{\bJ}{\bm{J}}
\newcommand{\bK}{\bm{K}}
\newcommand{\bA}{\bm{A}}
\newcommand{\bB}{\bm{B}}
\newcommand{\bP}{\bm{P}}
\newcommand{\bQ}{\bm{Q}}
\newcommand{\bR}{\bm{R}}
\newcommand{\bI}{\bm{I}}
\newcommand{\bzero}{\bm{0}}

\title{%
    \textbf{INDI Flight Controller for a Servoless Asymmetric Tricopter} \\[6pt]
    \large Incremental Nonlinear Dynamic Inversion + Priority Allocator + Altitude PID \\[4pt]
    \large Complete Mathematical Derivation, Software Architecture, \\ and Implementation Guide
}
\author{Auto-generated Technical Report}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

% =============================================================================
\section{Introduction and Motivation}
% =============================================================================

\subsection{What Is a Tricopter?}

A \textbf{tricopter} is a type of multirotor aircraft that uses exactly three propellers (rotors) to fly.  Each propeller is driven by a brushless DC motor.  Compared with the more common quadcopter (four rotors), a tricopter has one fewer actuator.  This makes it mechanically simpler but significantly harder to control.

In a quadcopter the four motors are typically arranged symmetrically and alternate in spin direction (clockwise / counter-clockwise), so the reaction torques cancel perfectly at equal speed.  With only three rotors, the reaction torques \emph{cannot} cancel in general, and a traditional tricopter uses a \textbf{yaw servo} --- a small motor that tilts the rear propeller --- to generate the missing yaw torque.

Our vehicle is a \textbf{servoless} tricopter: it has no yaw servo.  Yaw control must come entirely from the \emph{differential reaction torques} of the three fixed-spin propellers.  Furthermore, the rotor positions are \textbf{asymmetric} (the triangle formed by the three rotors is \emph{not} equilateral), and the full $3\times3$ inertia tensor has off-diagonal coupling terms.

\subsection{Why INDI?}

A previous version of this controller used a \textbf{Linear Quadratic Regulator (LQR)} to stabilise roll, pitch, and yaw.  The LQR linearises the dynamics about hover, designs an optimal gain matrix, and applies linear feedback.  While mathematically elegant, this approach has significant drawbacks for a servoless tricopter:

\begin{itemize}
    \item \textbf{Model dependency.}  The LQR requires an accurate linear model of the plant.  Any mismatch between the model and reality (unmodelled aerodynamics, mass changes, propeller wear) degrades performance.
    \item \textbf{Poor disturbance rejection.}  External torques (wind gusts, payload shifts) are not part of the linear model and can only be rejected through the slow feedback loop.
    \item \textbf{Yaw coupling.}  The LQR's optimal gain matrix couples all axes.  For a vehicle with 30$\times$ weaker yaw authority, this means yaw errors contaminate roll/pitch control.
\end{itemize}

\textbf{Incremental Nonlinear Dynamic Inversion (INDI)} solves all three problems.  Instead of using a model to predict what the motors \emph{should} produce, INDI \emph{measures} what the motors \emph{actually} produced (via a filtered angular acceleration estimate) and computes only the \emph{incremental change} needed.  This means:

\begin{enumerate}[label=\arabic*.]
    \item \textbf{Model errors cancel.}  The only model knowledge INDI needs is the \emph{control effectiveness matrix} $\bm{G}$ --- how motor speed changes map to angular acceleration changes.  Gravity, aerodynamic drag, gyroscopic coupling, and all other nonlinear terms cancel in the incremental formulation.
    \item \textbf{Natural disturbance rejection.}  External torques appear in the measured angular acceleration and are automatically compensated at the next timestep.
    \item \textbf{Clean axis separation.}  INDI computes the desired angular acceleration for each axis independently, then uses the inverse of $\bm{G}$ to find the required motor changes.
\end{enumerate}

Our INDI architecture has four components:
\begin{enumerate}[label=\arabic*.]
    \item \textbf{INDI Controller (inner, fast):} a 3-loop cascaded controller (attitude P $\to$ rate P $\to$ incremental inversion) that computes desired motor speed changes from attitude and rate errors.
    \item \textbf{Priority-Based Allocator (inner):} applies the motor changes while respecting actuator limits, sacrificing yaw authority before roll/pitch when motors saturate.
    \item \textbf{Altitude PID (outer, slow):} a classical PID controller that regulates altitude by commanding total thrust.
    \item \textbf{Heading-Aware Command (infrastructure):} rotates future world-frame commands through the current yaw angle, enabling correct position control despite yaw drift.
\end{enumerate}
Each timestep: the INDI controller computes attitude corrections, the priority allocator applies them within actuator limits, the altitude PID adjusts collective thrust as a post-allocation correction, and the motor model filters the commanded speeds through a first-order lag.

\subsection{Control Input Parameterisation}

A critical design choice: we use $\omega_i^2$ (the \emph{square} of each motor's angular velocity) as the control variable, \textbf{not} thrust directly.  Why?  Because both the thrust force and the reaction drag torque of a propeller are \emph{linear} in $\omega^2$:
\begin{align}
    T_i &= k_{T,i} \, \omega_i^2  \label{eq:thrust} \\
    Q_i &= k_{Q,i} \, \omega_i^2  \label{eq:drag}
\end{align}
where $k_T$ is the thrust coefficient and $k_Q$ is the drag torque coefficient of the propeller.  Because both are linear in $\omega^2$, the mapping from the control vector $\bu = [\omega_1^2, \omega_2^2, \omega_3^2]^T$ to forces and torques is a \emph{linear matrix equation}.  This linearity is essential for INDI: the control effectiveness matrix $\bm{G}$ (which maps $\Delta\bu$ to $\Delta\dot{\bomega}$) is constant, making the incremental inversion exact.

% =============================================================================
\section{Vehicle Model}
% =============================================================================

\subsection{Coordinate Frames}

We use two coordinate frames:
\begin{itemize}
    \item \textbf{World frame} (inertial): $x$ forward, $y$ left, $z$ up.  Gravity points in the $-z$ direction: $\bm{g} = [0, 0, -9.81]^T$ m/s$^2$.
    \item \textbf{Body frame}: fixed to the vehicle's centre of gravity (CG).  Axes align with the vehicle's principal directions.  All rotor positions, thrust axes, and spin axes are specified in this frame.
\end{itemize}

\subsection{Rotor Configuration}

Each rotor $i$ (where $i = 1, 2, 3$) is described by:
\begin{itemize}
    \item $\bm{r}_i \in \R^3$: position vector from the CG to the rotor, in body frame [m].
    \item $\hat{\bm{t}}_i \in \R^3$: unit thrust axis (direction of thrust force).
    \item $\hat{\bm{s}}_i \in \R^3$: unit spin axis (axis about which the propeller rotates).
    \item $d_i \in \{+1, -1\}$: spin direction ($+1$ = clockwise, $-1$ = counter-clockwise when viewed from above).
    \item $k_{T,i}$: thrust coefficient [N$\cdot$s$^2$/rad$^2$].
    \item $k_{Q,i}$: drag torque coefficient [N$\cdot$m$\cdot$s$^2$/rad$^2$].
\end{itemize}

For our default configuration:
\begin{table}[H]
\centering
\begin{tabular}{lccccc}
\toprule
\textbf{Rotor} & \textbf{Position} [m] & $\hat{\bm{t}}_i$ & $d_i$ & $k_T$ & $k_Q$ \\
\midrule
0 (front)       & $(0.25, 0, 0)$       & $(0,0,1)$ & $+1$ & $1.5\times10^{-5}$ & $2.5\times10^{-7}$ \\
1 (rear-left)   & $(-0.15, 0.20, 0)$   & $(0,0,1)$ & $-1$ & $1.5\times10^{-5}$ & $2.5\times10^{-7}$ \\
2 (rear-right)  & $(-0.15, -0.20, 0)$  & $(0,0,1)$ & $+1$ & $1.5\times10^{-5}$ & $2.5\times10^{-7}$ \\
\bottomrule
\end{tabular}
\caption{Default rotor configuration.  Note the asymmetric positions and two CW vs one CCW spin direction.}
\label{tab:rotors}
\end{table}

\subsection{Inertia Tensor}

The vehicle has mass $m = 1.5$ kg and a \emph{full} (non-diagonal) inertia tensor:
\begin{equation}
    \bJ = \begin{bmatrix} J_{xx} & J_{xy} & J_{xz} \\ J_{xy} & J_{yy} & J_{yz} \\ J_{xz} & J_{yz} & J_{zz} \end{bmatrix}
    = \begin{bmatrix} 0.03 & 0.001 & 0.0005 \\ 0.001 & 0.025 & 0.0008 \\ 0.0005 & 0.0008 & 0.04 \end{bmatrix}
    \; \text{kg}\cdot\text{m}^2
\end{equation}
The off-diagonal terms arise because the vehicle is not symmetric about any axis.  A diagonal inertia tensor would imply perfect symmetry.

% =============================================================================
\section{Control Allocation}
\label{sec:control_alloc}
% =============================================================================

\subsection{The Allocation Matrix}

The \textbf{control allocation matrix} is the linear map from motor commands $\bu = [\omega_1^2, \omega_2^2, \omega_3^2]^T$ to the total body-frame forces and torques:
\begin{equation}
    \begin{bmatrix} \bF_{\text{body}} \\ \btau_{\text{body}} \end{bmatrix}
    = \bB_{\text{alloc}} \, \bu
    \qquad
    \bB_{\text{alloc}} \in \R^{6 \times 3}
\end{equation}

Each column of $\bB_{\text{alloc}}$ represents the force and torque produced by rotor $i$ per unit $\omega_i^2$.

\subsubsection{Force Column}

The force from rotor $i$ is simply thrust along its thrust axis:
\begin{equation}
    \bm{f}_i = k_{T,i} \, \hat{\bm{t}}_i
\end{equation}
For our rotors with $\hat{\bm{t}}_i = [0, 0, 1]^T$ and $k_T = 1.5 \times 10^{-5}$, every force column is $[0, 0, 1.5 \times 10^{-5}]^T$.

\subsubsection{Torque Column}

The torque from rotor $i$ has \textbf{two components}:

\paragraph{1. Moment from thrust (cross-product torque).}  The thrust force $\bm{f}_i$ acts at position $\bm{r}_i$ from the CG, creating a moment:
\begin{equation}
    \btau_{\text{thrust},i} = \bm{r}_i \times (k_{T,i} \, \hat{\bm{t}}_i)
\end{equation}
This is the familiar ``force times lever arm'' torque.  For example, a rotor offset in the $y$-direction creates a roll torque (about the $x$-axis).

\paragraph{2. Reaction (drag) torque.}  As the propeller spins, aerodynamic drag on the blades produces a torque that opposes the rotation.  By Newton's third law, this reaction torque is applied to the vehicle:
\begin{equation}
    \btau_{\text{drag},i} = d_i \, k_{Q,i} \, \hat{\bm{s}}_i
\end{equation}
The sign $d_i = \pm 1$ indicates the spin direction.  A clockwise-spinning propeller ($d_i = +1$) creates a positive torque about the spin axis on the vehicle body.

\paragraph{Total torque column:}
\begin{equation}
    \btau_i = \bm{r}_i \times (k_{T,i} \, \hat{\bm{t}}_i) + d_i \, k_{Q,i} \, \hat{\bm{s}}_i
    \label{eq:torque_col}
\end{equation}

\subsubsection{Worked Example: Rotor 0 (Front)}

\begin{align}
    \bm{f}_0 &= 1.5 \times 10^{-5} \cdot [0, 0, 1]^T = [0, 0, 1.5 \times 10^{-5}]^T \\
    \btau_{\text{thrust},0} &= [0.25, 0, 0]^T \times [0, 0, 1.5 \times 10^{-5}]^T \\
    &= \begin{bmatrix}
        0 \cdot 1.5\times10^{-5} - 0 \cdot 0 \\
        0 \cdot 0 - 0.25 \cdot 1.5\times10^{-5} \\
        0.25 \cdot 0 - 0 \cdot 0
    \end{bmatrix}
    = \begin{bmatrix} 0 \\ -3.75 \times 10^{-6} \\ 0 \end{bmatrix} \\
    \btau_{\text{drag},0} &= (+1) \cdot 2.5 \times 10^{-7} \cdot [0, 0, 1]^T = [0, 0, 2.5 \times 10^{-7}]^T \\
    \btau_0 &= [0, \; -3.75 \times 10^{-6}, \; 2.5 \times 10^{-7}]^T
\end{align}

The front rotor creates \emph{negative} pitch torque (nose-down tendency due to its forward position) and a small positive yaw torque (reaction from CW spin).

\subsection{The Full Allocation Matrix}

Assembling all columns:
\begin{equation}
    \bB_{\text{alloc}} = \begin{bmatrix}
        0 & 0 & 0 \\
        0 & 0 & 0 \\
        1.5\times10^{-5} & 1.5\times10^{-5} & 1.5\times10^{-5} \\
        0 & 3.0\times10^{-6} & -3.0\times10^{-6} \\
        -3.75\times10^{-6} & 2.25\times10^{-6} & 2.25\times10^{-6} \\
        2.5\times10^{-7} & -2.5\times10^{-7} & 2.5\times10^{-7}
    \end{bmatrix}
\end{equation}
Rows 1--3 are force ($\bB_{\text{force}}$), rows 4--6 are torque ($\bB_\tau$).

\subsection{Controllability Analysis of the Torque Sub-Block}

The torque sub-block $\bB_\tau \in \R^{3 \times 3}$ tells us how well we can control the three torque axes (roll, pitch, yaw):
\begin{equation}
    \bB_\tau = \begin{bmatrix}
        0 & 3.0\times10^{-6} & -3.0\times10^{-6} \\
        -3.75\times10^{-6} & 2.25\times10^{-6} & 2.25\times10^{-6} \\
        2.5\times10^{-7} & -2.5\times10^{-7} & 2.5\times10^{-7}
    \end{bmatrix}
\end{equation}

The \textbf{singular value decomposition} (SVD) reveals:
\begin{align}
    \sigma_1 &= 4.92 \times 10^{-6} \quad \text{(strongest axis)} \\
    \sigma_2 &= 4.26 \times 10^{-6} \\
    \sigma_3 &= 1.61 \times 10^{-7} \quad \text{(weakest axis --- yaw)}
\end{align}

The \textbf{condition number} $\kappa = \sigma_1 / \sigma_3 = 30.6$.  This means the yaw axis is about 30 times weaker than the strongest torque axis.  The \textbf{rank} is 3 (full rank), meaning all three torque axes \emph{are} controllable, but yaw has very limited authority.

\subsection{Physical Interpretation}

Roll and pitch are controlled by the large lever arms of the rotor positions (up to $\bm{r}_i = 0.25$ m), multiplied by $k_T = 1.5 \times 10^{-5}$, giving torque sensitivity $\sim 3.75 \times 10^{-6}$ N$\cdot$m per unit $\omega^2$.

Yaw is controlled \emph{only} by the tiny drag torque coefficient $k_Q = 2.5 \times 10^{-7}$, which is about 15 times smaller.  This weak yaw authority is a fundamental property of servoless tricopters.

% =============================================================================
\section{Hover Trim}
\label{sec:trim}
% =============================================================================

\subsection{The Trim Problem}

``Trim'' means finding the motor speeds that produce \textbf{steady hover}: the vehicle hangs motionless in the air.  This requires:
\begin{enumerate}
    \item \textbf{Force balance:} Total thrust equals weight.
    \begin{equation}
        \sum_{i=1}^{3} k_{T,i} \, (\hat{\bm{t}}_i \cdot \hat{\bm{z}}) \, u_i = m g
        \label{eq:thrust_balance}
    \end{equation}
    where $u_i = \omega_i^2$ and $\hat{\bm{z}} = [0,0,1]^T$.

    \item \textbf{Torque balance:} Net torque is zero.
    \begin{equation}
        \bB_\tau \, \bu = \bm{0}
        \label{eq:torque_balance}
    \end{equation}
\end{enumerate}

\subsection{Why Exact Trim Is Impossible}

Equation~\eqref{eq:torque_balance} is 3 equations in 3 unknowns.  Equation~\eqref{eq:thrust_balance} adds a 4th equation.  Together we have \textbf{4 equations in 3 unknowns} --- an \emph{overdetermined} system.

Since $\bB_\tau$ is full rank (rank 3), the only solution to $\bB_\tau \bu = \bm{0}$ is $\bu = \bm{0}$, which gives zero thrust.  Therefore, \textbf{no motor speed combination can simultaneously produce zero torque and nonzero thrust} for this vehicle geometry.

This is a fundamental consequence of having 3 actuators and 4 constraints.  Conventional tricopters solve this with a yaw servo (4th actuator).

\subsection{The KKT Optimisation}

Since we cannot satisfy all 4 constraints exactly, we find the \emph{best compromise}: minimise the residual torque while exactly meeting the thrust requirement.  This is a \textbf{constrained quadratic optimisation}:
\begin{equation}
    \min_{\bu} \; \bu^T \underbrace{\bB_\tau^T \bB_\tau}_{\bm{M}} \bu
    \qquad \text{subject to} \quad \bm{c}^T \bu = mg
\end{equation}
where $\bm{c} = [k_{T,1}, k_{T,2}, k_{T,3}]^T$ is the thrust coefficient vector (assuming $\hat{\bm{t}}_i \cdot \hat{\bm{z}} = 1$).

The \textbf{Karush-Kuhn-Tucker (KKT)} conditions for this problem are obtained by forming the Lagrangian:
\begin{equation}
    \mathcal{L}(\bu, \lambda) = \bu^T \bm{M} \bu + \lambda \, (\bm{c}^T \bu - mg)
\end{equation}
Setting $\nabla_{\bu} \mathcal{L} = 0$ and $\nabla_\lambda \mathcal{L} = 0$:
\begin{equation}
    \begin{bmatrix} 2\bm{M} & \bm{c} \\ \bm{c}^T & 0 \end{bmatrix}
    \begin{bmatrix} \bu \\ \lambda \end{bmatrix}
    = \begin{bmatrix} \bm{0} \\ mg \end{bmatrix}
    \label{eq:kkt}
\end{equation}
This is a $(N{+}1) \times (N{+}1)$ linear system (4$\times$4 for our tricopter) that we solve directly.

\subsection{Trim Results}

The KKT solver produces:
\begin{align}
    \omega_0^2 &= 367{,}242 \quad \Rightarrow \quad \omega_0 = 606.0 \;\text{rad/s} \quad \text{(front)} \\
    \omega_1^2 &= 308{,}145 \quad \Rightarrow \quad \omega_1 = 555.1 \;\text{rad/s} \quad \text{(rear-left)} \\
    \omega_2^2 &= 305{,}613 \quad \Rightarrow \quad \omega_2 = 552.8 \;\text{rad/s} \quad \text{(rear-right)}
\end{align}
The speeds are \textbf{unequal}, as expected.  The front rotor spins fastest because:
\begin{itemize}
    \item It must produce the most thrust for pitch balance (its lever arm is largest: $x = 0.25$ m vs $0.15$ m for the rear pair).
    \item Being the sole CW rotor at the front, it carries extra yaw-balancing burden.
\end{itemize}

The residual torque is $[0.0076, \; 0.0038, \; 0.0912]^T$ Nm, with most of the residual in yaw (0.091 Nm).  This is the unavoidable physical limitation discussed above.

% =============================================================================
\section{Rigid Body Dynamics}
\label{sec:dynamics}
% =============================================================================

\subsection{State Vector}

The full nonlinear state vector has 13 components:
\begin{equation}
    \bx = \begin{bmatrix}
        p_x \\ p_y \\ p_z \\ v_x \\ v_y \\ v_z \\ q_w \\ q_x \\ q_y \\ q_z \\ \omega_x \\ \omega_y \\ \omega_z
    \end{bmatrix}
    \in \R^{13}
\end{equation}
where:
\begin{itemize}
    \item $\bm{p} = [p_x, p_y, p_z]^T$: position in world frame [m].
    \item $\bm{v} = [v_x, v_y, v_z]^T$: velocity in world frame [m/s].
    \item $\bq = [q_w, q_x, q_y, q_z]^T$: unit quaternion representing body-to-world rotation.
    \item $\bomega = [\omega_x, \omega_y, \omega_z]^T$: angular velocity in body frame [rad/s].
\end{itemize}

\subsection{Why Quaternions?}

We represent attitude using \textbf{quaternions} rather than Euler angles because:
\begin{enumerate}
    \item \textbf{No gimbal lock.}  Euler angles have a singularity at $\pm 90°$ pitch where yaw and roll become indistinguishable.  Quaternions are singularity-free.
    \item \textbf{Smooth integration.}  Quaternion kinematics are a simple linear ODE in $\bomega$, making numerical integration stable.
    \item \textbf{Efficient rotation.}  Converting a quaternion to a rotation matrix requires only 9 multiplications and 6 additions.
\end{enumerate}

A unit quaternion $\bq = [q_w, q_x, q_y, q_z]^T$ with $\|\bq\| = 1$ represents a rotation.  The scalar part $q_w$ encodes the half-angle of rotation, and the vector part $[q_x, q_y, q_z]$ encodes the axis.

The rotation matrix from body to world frame is:
\begin{equation}
    \bm{R}(\bq) = \begin{bmatrix}
        1 - 2(q_y^2 + q_z^2) & 2(q_x q_y - q_w q_z) & 2(q_x q_z + q_w q_y) \\
        2(q_x q_y + q_w q_z) & 1 - 2(q_x^2 + q_z^2) & 2(q_y q_z - q_w q_x) \\
        2(q_x q_z - q_w q_y) & 2(q_y q_z + q_w q_x) & 1 - 2(q_x^2 + q_y^2)
    \end{bmatrix}
\end{equation}

\subsection{Translational Dynamics}

Newton's second law in the world frame:
\begin{equation}
    m \, \dot{\bm{v}} = \bm{R}(\bq) \, \bF_{\text{body}} + m \, \bm{g}
    \label{eq:trans_dyn}
\end{equation}
where $\bF_{\text{body}} = \bB_{\text{force}} \, \bu$ is the total force in the body frame and $\bm{g} = [0, 0, -9.81]^T$ m/s$^2$.

The rotation matrix $\bm{R}(\bq)$ transforms the body-frame force into the world frame.  At hover ($\bq = [1,0,0,0]$, i.e.\ identity rotation), $\bm{R} = \bI$ and the thrust points straight up.  When the vehicle tilts, $\bm{R}$ redirects some thrust sideways.

\subsection{Rotational Dynamics (Euler's Equation)}

The angular momentum equation in the body frame:
\begin{equation}
    \bJ \, \dot{\bomega} = \btau_{\text{total}} - \bomega \times (\bJ \, \bomega)
    \label{eq:euler_eq}
\end{equation}

Here:
\begin{itemize}
    \item $\bJ$ is the $3 \times 3$ inertia tensor.
    \item $\btau_{\text{total}} = \bB_\tau \, \bu + \btau_{\text{external}}$ is the net torque (from motors plus any external disturbance).
    \item $\bomega \times (\bJ \bomega)$ is the \textbf{gyroscopic coupling} term.  This is zero at hover ($\bomega = \bm{0}$) but becomes important during manoeuvres.
\end{itemize}

Solving for $\dot{\bomega}$:
\begin{equation}
    \dot{\bomega} = \bJ^{-1} \left( \btau_{\text{total}} - \bomega \times (\bJ \bomega) \right)
\end{equation}

Note: because $\bJ$ has off-diagonal terms, $\bJ^{-1}$ couples all three axes.  An angular acceleration about one axis causes angular acceleration about the others.

\subsection{Quaternion Kinematics}

The time derivative of the quaternion:
\begin{equation}
    \dot{\bq} = \frac{1}{2} \, \bq \otimes \begin{bmatrix} 0 \\ \bomega \end{bmatrix}
    \label{eq:quat_kin}
\end{equation}
where $\otimes$ is the Hamilton product (quaternion multiplication).  Expanding:
\begin{align}
    \dot{q}_w &= \tfrac{1}{2} (-q_x \omega_x - q_y \omega_y - q_z \omega_z) \\
    \dot{q}_x &= \tfrac{1}{2} (+q_w \omega_x + q_y \omega_z - q_z \omega_y) \\
    \dot{q}_y &= \tfrac{1}{2} (+q_w \omega_y - q_x \omega_z + q_z \omega_x) \\
    \dot{q}_z &= \tfrac{1}{2} (+q_w \omega_z + q_x \omega_y - q_y \omega_x)
\end{align}

After each integration step, the quaternion must be \textbf{renormalised} to $\|\bq\| = 1$ to prevent numerical drift from the unit sphere.

\subsection{Euler Angle Extraction (ZYX Convention)}

For logging and for the INDI attitude error computation, we extract Euler angles from the quaternion using the ZYX (yaw-pitch-roll) convention:
\begin{align}
    \phi \;\;\text{(roll)} &= \text{atan2}(2(q_w q_x + q_y q_z), \; 1 - 2(q_x^2 + q_y^2)) \\
    \theta \;\text{(pitch)} &= \arcsin(2(q_w q_y - q_z q_x)) \\
    \psi \;\;\text{(yaw)} &= \text{atan2}(2(q_w q_z + q_x q_y), \; 1 - 2(q_y^2 + q_z^2))
\end{align}
These are used \emph{only} for display and as the error signal to the INDI attitude controller.  All internal dynamics use quaternions.

% =============================================================================
\section{Numerical Integration: 4th-Order Runge-Kutta}
\label{sec:rk4}
% =============================================================================

\subsection{Why RK4?}

We need to advance the 13-state ODE $\dot{\bx} = \bm{f}(\bx, \bu)$ by a small timestep $\Delta t$.  The simplest method (forward Euler: $\bx_{k+1} = \bx_k + \Delta t \, \bm{f}(\bx_k, \bu_k)$) is only first-order accurate and requires very small $\Delta t$ for stability.

The \textbf{classical 4th-order Runge-Kutta} (RK4) method evaluates the derivative at 4 points within each timestep and takes a weighted average:
\begin{align}
    \bm{k}_1 &= \bm{f}(\bx_k, \bu_k) \\
    \bm{k}_2 &= \bm{f}\!\left(\bx_k + \tfrac{\Delta t}{2} \bm{k}_1, \; \bu_k\right) \\
    \bm{k}_3 &= \bm{f}\!\left(\bx_k + \tfrac{\Delta t}{2} \bm{k}_2, \; \bu_k\right) \\
    \bm{k}_4 &= \bm{f}(\bx_k + \Delta t \, \bm{k}_3, \; \bu_k) \\[4pt]
    \bx_{k+1} &= \bx_k + \frac{\Delta t}{6} \left( \bm{k}_1 + 2\bm{k}_2 + 2\bm{k}_3 + \bm{k}_4 \right)
\end{align}

The error per step is $\mathcal{O}(\Delta t^5)$, and the global error over the simulation is $\mathcal{O}(\Delta t^4)$.  With $\Delta t = 0.001$ s (1 kHz), RK4 provides excellent accuracy.

\subsection{Quaternion Renormalisation}

After the RK4 update, the quaternion may drift slightly from unit norm due to floating-point arithmetic.  We renormalise:
\begin{equation}
    \bq \leftarrow \frac{\bq}{\|\bq\|}
\end{equation}
This is done \emph{once} after the final RK4 update (not at intermediate stages), which is sufficient for maintaining $\|\bq\| = 1$ to machine precision.

% =============================================================================
\section{Inner Loop: INDI Controller}
\label{sec:indi}
% =============================================================================

\subsection{What Is INDI?}

\textbf{Incremental Nonlinear Dynamic Inversion (INDI)} is a sensor-based control technique.  Unlike classical controllers (PID, LQR) that use a full plant model, INDI uses \emph{measurements} of the current angular acceleration to cancel most of the dynamics, then applies only an \emph{incremental} correction.

The core idea is best understood by starting from Euler's rotational equation:
\begin{equation}
    \bJ \, \dot{\bomega} = \btau_{\text{motor}} + \btau_{\text{external}} - \bomega \times (\bJ \bomega)
    \label{eq:euler_indi}
\end{equation}

In a traditional controller, we would need to model every term on the right-hand side: motor torques, external disturbances, and gyroscopic coupling.  INDI takes a different approach.  At the current timestep, \emph{all} of those terms have already produced a specific angular acceleration $\dot{\bomega}_0$ that we can \emph{measure} (or estimate from gyroscope data).  If we want to change the angular acceleration by a small amount $\Delta\dot{\bomega}$, we only need to change the motor commands by a corresponding small amount $\Delta\bu$.

Writing Equation~\eqref{eq:euler_indi} at two successive instants and subtracting:
\begin{equation}
    \bJ \, (\dot{\bomega}_{\text{new}} - \dot{\bomega}_0) = \bB_\tau \, (\bu_{\text{new}} - \bu_0) + \underbrace{\Delta\btau_{\text{ext}} - \Delta(\bomega \times \bJ\bomega)}_{\approx \, 0 \text{ at high sample rate}}
\end{equation}

At a sufficiently high sample rate (1 kHz), the external torques and gyroscopic terms change negligibly between two consecutive timesteps, so the crossed-out terms vanish.  This gives the \textbf{incremental relation}:
\begin{equation}
    \boxed{
    \Delta\dot{\bomega} = \bJ^{-1} \bB_\tau \, \Delta\bu = \bm{G} \, \Delta\bu
    }
    \label{eq:indi_core}
\end{equation}

where $\bm{G} = \bJ^{-1} \bB_\tau \in \R^{3 \times 3}$ is the \textbf{control effectiveness matrix}.  This is the \emph{only} model knowledge INDI requires.

\subsection{The G Matrix: Control Effectiveness}

The control effectiveness matrix $\bm{G}$ maps changes in motor commands ($\Delta\bu = \Delta[\omega_1^2, \omega_2^2, \omega_3^2]^T$) to changes in angular acceleration ($\Delta\dot{\bomega} = \Delta[\dot{p}, \dot{q}, \dot{r}]^T$):

\begin{equation}
    \bm{G} = \bJ^{-1} \bB_\tau
\end{equation}

For our vehicle:
\begin{equation}
    \bm{G} = \begin{bmatrix}
        4.86 \times 10^{-6} & 9.73 \times 10^{-5} & -1.03 \times 10^{-4} \\
        -1.50 \times 10^{-4} & 8.64 \times 10^{-5} & 9.39 \times 10^{-5} \\
        9.20 \times 10^{-6} & -9.19 \times 10^{-6} & 5.66 \times 10^{-6}
    \end{bmatrix}
\end{equation}

Reading this matrix:
\begin{itemize}
    \item \textbf{Row 0} (roll acceleration $\dot{p}$): Most sensitive to motors 1 and 2 (rear pair, which have $y$-axis lever arms for roll torque).  Motor 0 (front, on the x-axis) has almost no roll effect.
    \item \textbf{Row 1} (pitch acceleration $\dot{q}$): Sensitive to all three motors, with motor 0 (large forward arm) having the strongest effect.
    \item \textbf{Row 2} (yaw acceleration $\dot{r}$): All entries are very small (order $10^{-6}$ vs $10^{-4}$ for roll/pitch).  This reflects the fundamental weakness of yaw control via reaction torques alone.
\end{itemize}

\subsubsection{G Matrix Condition Number}

The condition number of $\bm{G}$, computed via singular value decomposition (SVD), is:
\begin{equation}
    \kappa(\bm{G}) = \frac{\sigma_{\max}}{\sigma_{\min}} = 49.25
\end{equation}

This means the strongest control axis is about 49 times more effective than the weakest (yaw).  When we invert $\bm{G}$ to compute $\bm{G}^{-1}$, the weak yaw axis gets amplified by a factor of $\sim 49$, producing very large entries in the yaw column of $\bm{G}^{-1}$.

\subsubsection{The Inverse: $\bm{G}^{-1}$}

Since $\bm{G}$ is $3 \times 3$ and full rank, we can invert it directly:
\begin{equation}
    \Delta\bu = \bm{G}^{-1} \, \Delta\dot{\bomega}_{\text{des}}
\end{equation}

The entries of $\bm{G}^{-1}$ are of order $10^3$--$10^5$.  The yaw column (column 2) has particularly large entries ($\sim 1.3 \times 10^5$ to $1.6 \times 10^5$), meaning that requesting even a small yaw acceleration change requires enormous motor speed changes.  This is the mathematical manifestation of the weak yaw authority.

\subsection{The Three Cascaded Loops}

INDI uses three nested feedback loops, each operating on a successively faster signal:

\subsubsection{Loop 1: Attitude Proportional Controller}

The outermost loop converts attitude errors into desired angular rate commands:
\begin{align}
    \omega_{x,\text{cmd}} &= K_{p,\phi} \cdot (\phi_{\text{cmd}} - \phi) \\
    \omega_{y,\text{cmd}} &= K_{p,\theta} \cdot (\theta_{\text{cmd}} - \theta) \\
    \omega_{z,\text{cmd}} &= -K_{p,\psi} \cdot r
\end{align}

For roll and pitch, the controller drives the angle error to zero.  For yaw, it acts as a \textbf{rate damper} only: it does not try to hold a heading (which is impossible for this vehicle), but instead opposes the current yaw rate $r$ to prevent unbounded spinning.

Default gains: $K_{p,\phi} = K_{p,\theta} = 6.0$, $K_{p,\psi} = 2.0$.

\subsubsection{Loop 2: Rate Proportional Controller}

The middle loop converts angular rate errors into desired angular acceleration commands:
\begin{equation}
    \dot{\bomega}_{\text{cmd}} = \bm{K}_{p,\text{rate}} \circ (\bomega_{\text{cmd}} - \bomega)
\end{equation}

where $\circ$ denotes element-wise multiplication and $\bm{K}_{p,\text{rate}} = [15, 15, 8]^T$.

This is analogous to a derivative term: it provides damping to prevent oscillation in the attitude response.

\subsubsection{Loop 3: Incremental Inversion (the INDI step)}

The innermost loop computes the actual motor command change:
\begin{equation}
    \boxed{
    \Delta\bu = \bm{G}^{-1} \left( \dot{\bomega}_{\text{cmd}} - \dot{\bomega}_{\text{filtered}} \right)
    }
    \label{eq:indi_law}
\end{equation}

Here $\dot{\bomega}_{\text{filtered}}$ is the \emph{measured} angular acceleration, estimated by differentiating the gyroscope signal and passing it through a low-pass filter (Section~\ref{sec:butterworth}).  The difference $\dot{\bomega}_{\text{cmd}} - \dot{\bomega}_{\text{filtered}}$ is the \emph{incremental change} in angular acceleration that we want.

This is the key equation.  It says: ``I am currently experiencing angular acceleration $\dot{\bomega}_{\text{filtered}}$.  I want angular acceleration $\dot{\bomega}_{\text{cmd}}$.  The difference is $\Delta\dot{\bomega}$, and $\bm{G}^{-1}$ tells me which motor changes produce that.''

\subsection{Why INDI Is Robust}

The power of INDI comes from the cancellation in Equation~\eqref{eq:indi_law}.  Consider what happens when an external wind gust applies a torque:

\begin{enumerate}
    \item The gust creates an unexpected angular acceleration, which appears in $\dot{\bomega}_{\text{filtered}}$.
    \item The desired angular acceleration $\dot{\bomega}_{\text{cmd}}$ has not changed (the attitude/rate controllers still want to hover).
    \item The difference $\dot{\bomega}_{\text{cmd}} - \dot{\bomega}_{\text{filtered}}$ automatically includes a correction for the gust.
    \item $\bm{G}^{-1}$ translates this into motor commands that counteract the gust.
\end{enumerate}

This happens within a single control loop iteration --- no model of the wind is needed.  Similarly, if the vehicle's mass changes (e.g., payload added), the gyroscopic coupling term $\bomega \times (\bJ\bomega)$ changes, but INDI compensates automatically because the change appears in the measured $\dot{\bomega}_{\text{filtered}}$.

The only error source is if $\bm{G}$ itself is wrong (e.g., a motor loses thrust).  Even then, INDI degrades gracefully because the feedback loop corrects for the error at the next timestep.

% =============================================================================
\section{Angular Acceleration Estimation: Butterworth Filter}
\label{sec:butterworth}
% =============================================================================

\subsection{The Estimation Problem}

INDI requires the angular acceleration $\dot{\bomega}$, but gyroscopes measure angular \emph{rate} $\bomega$, not its derivative.  The simplest approach --- finite differencing $\dot{\bomega} \approx (\bomega_k - \bomega_{k-1})/\Delta t$ --- amplifies high-frequency noise catastrophically.  At 1 kHz sampling, even small gyroscope noise ($\sim 0.01$ rad/s RMS) produces acceleration noise of $\sim 10$ rad/s$^2$.

The solution is to low-pass filter the differentiated signal.  We use a \textbf{2nd-order Butterworth} filter because:
\begin{enumerate}
    \item It has the \textbf{flattest possible passband} response (maximally flat magnitude).  This means it does not distort the signal below the cutoff frequency.
    \item It provides $-40$ dB/decade roll-off above the cutoff, which attenuates noise efficiently.
    \item It has well-understood phase characteristics (smooth, predictable phase lag).
\end{enumerate}

\subsection{Butterworth Filter Design}

The continuous-time transfer function of a 2nd-order Butterworth low-pass filter is:
\begin{equation}
    H(s) = \frac{\omega_c^2}{s^2 + \sqrt{2}\,\omega_c\,s + \omega_c^2}
\end{equation}
where $\omega_c = 2\pi f_c$ is the cutoff frequency in rad/s.

To implement this digitally, we use the \textbf{bilinear transform} (Tustin's method) with \textbf{frequency pre-warping}:

\subsubsection{Step 1: Pre-warped frequency}

The bilinear transform introduces frequency warping.  To ensure the digital filter's $-3$ dB point matches the desired analog cutoff $f_c$, we compute the pre-warped discrete frequency:
\begin{equation}
    \gamma = \tan\!\left(\frac{\pi \, f_c}{f_s}\right)
\end{equation}
where $f_s$ is the sample rate (1000 Hz) and $f_c$ is the cutoff frequency (30 Hz).

For our parameters: $\gamma = \tan(\pi \cdot 30 / 1000) = \tan(0.09425) = 0.09480$.

\subsubsection{Step 2: Coefficient computation}

The normalisation denominator is:
\begin{equation}
    D = \gamma^2 + \sqrt{2}\,\gamma + 1
\end{equation}

The filter coefficients are:
\begin{align}
    b_0 &= b_2 = \frac{\gamma^2}{D} \\
    b_1 &= \frac{2\gamma^2}{D} \\
    a_1 &= \frac{2(\gamma^2 - 1)}{D} \\
    a_2 &= \frac{\gamma^2 - \sqrt{2}\,\gamma + 1}{D}
\end{align}

For our parameters ($f_c = 30$ Hz, $f_s = 1000$ Hz):
\begin{align}
    b_0 &= b_2 = 0.00782 \\
    b_1 &= 0.01564 \\
    a_1 &= -1.7347 \\
    a_2 &= 0.7660
\end{align}

\subsubsection{Step 3: Difference equation}

The filter is implemented as the standard second-order IIR difference equation:
\begin{equation}
    y[n] = b_0 \, x[n] + b_1 \, x[n{-}1] + b_2 \, x[n{-}2] - a_1 \, y[n{-}1] - a_2 \, y[n{-}2]
\end{equation}

This requires storing only 4 values per axis: the previous two inputs $x[n{-}1], x[n{-}2]$ and the previous two outputs $y[n{-}1], y[n{-}2]$.  Three independent filter instances are used for roll, pitch, and yaw angular acceleration.

\subsection{Filter Characteristics}

The 2nd-order Butterworth with 30 Hz cutoff at 1 kHz sampling has these properties:
\begin{itemize}
    \item \textbf{DC gain:} Exactly 1.0 (the sum $b_0 + b_1 + b_2 = 0.03128$, divided by $1 + a_1 + a_2 = 0.03128$, equals 1).
    \item \textbf{$-3$ dB frequency:} 30 Hz (by construction).
    \item \textbf{Gain at cutoff:} $\approx 0.707$ ($= 1/\sqrt{2}$).
    \item \textbf{Phase lag at cutoff:} $-90°$ (characteristic of Butterworth filters).
    \item \textbf{Step response settling:} Reaches 99\% of final value within approximately 50 ms (1.5 time constants of the equivalent analog filter).
\end{itemize}

\subsection{Filter Pre-Seeding}

At simulation start (timestep 0), the filter states are zero, but the vehicle is already at hover trim with nonzero motor torques producing nonzero angular acceleration.  If the filter starts from zero, it takes $\sim$50 ms to ``warm up'' to the correct value.  During this warm-up, INDI sees $\dot{\bomega}_{\text{filtered}} \approx 0$ and thinks the vehicle has no angular acceleration, causing large incorrect motor commands.

The solution is to \textbf{pre-seed} the filter: set its internal states ($x[n{-}1], x[n{-}2], y[n{-}1], y[n{-}2]$) to the DC steady-state values corresponding to the trim angular acceleration:
\begin{equation}
    \dot{\bomega}_{\text{trim}} = \bJ^{-1} \bB_\tau \, \bu_{\text{hover}} = [0.213, \; 0.071, \; 2.275]^T \; \text{rad/s}^2
\end{equation}

\textbf{Critical detail:} We pre-seed \emph{only} the roll and pitch components.  The yaw component (2.275 rad/s$^2$) is deliberately set to zero in the pre-seed.  The reason is explained in Section~\ref{sec:thrust_collapse}: pre-seeding yaw causes catastrophic thrust collapse due to the large entries in $\bm{G}^{-1}$'s yaw column.

% =============================================================================
\section{Motor Dynamics Model}
\label{sec:motor_model}
% =============================================================================

\subsection{Why Model Motor Dynamics?}

Real brushless motors cannot change speed instantaneously.  The electrical and mechanical time constants create a lag between the commanded speed and the actual speed.  For INDI, this matters because:

\begin{enumerate}
    \item The motor's actual speed determines the actual torque, not the commanded speed.
    \item INDI's incremental law $\Delta\bu = \bm{G}^{-1}\Delta\dot{\bomega}$ assumes $\Delta\bu$ takes effect immediately.  Motor lag violates this assumption.
    \item Tracking $\bu_{\text{prev}} = \bu_{\text{actual}}$ (not $\bu_{\text{commanded}}$) is critical for correct incremental computation.
\end{enumerate}

\subsection{First-Order Lag Model}

We model each motor as a first-order lag:
\begin{equation}
    \dot{u}_{\text{actual},i} = \frac{u_{\text{cmd},i} - u_{\text{actual},i}}{\tau_m}
\end{equation}

where $\tau_m = 0.03$ s (30 ms) is the motor time constant.  Discretising with forward Euler at timestep $\Delta t$:
\begin{equation}
    u_{\text{actual},i}[k+1] = u_{\text{actual},i}[k] + \frac{\Delta t}{\tau_m} \left( u_{\text{cmd},i}[k] - u_{\text{actual},i}[k] \right)
\end{equation}

After updating, the output is clamped to $[0, \omega_{\max}^2]$ (motors cannot spin backwards or exceed maximum speed).

\subsection{Physical Interpretation}

At each timestep, the motor moves $\Delta t / \tau_m = 0.001 / 0.03 = 3.33\%$ of the way from its current speed to the commanded speed.  After one time constant ($\tau_m = 30$ ms $= 30$ timesteps), the motor has reached $1 - e^{-1} \approx 63.2\%$ of the commanded value.  After three time constants (90 ms), it reaches $1 - e^{-3} \approx 95\%$.

The 30 ms time constant is typical for small multirotor brushless motors (2204--2207 size) with ESCs running BLHeli\_S firmware.  Larger motors have longer time constants (50--100 ms); racing motors can be as fast as 10--15 ms.

\subsection{Why $\bu_{\text{prev}} = \bu_{\text{actual}}$ Is Critical}

INDI's incremental law computes $\Delta\bu$ relative to the \emph{previous} motor command.  If we track $\bu_{\text{prev}} = \bu_{\text{commanded}}$ (the ideal command), but the motors have not yet reached that command due to lag, then the incremental computation is wrong: it assumes the motors are somewhere they are not.

By tracking $\bu_{\text{prev}} = \bu_{\text{actual}}$ (the motor model's output), the incremental computation correctly accounts for motor lag.  The difference $\bu_{\text{new}} - \bu_{\text{actual}}$ represents the true change that needs to happen, not the ideal change.

% =============================================================================
\section{Priority-Based Control Allocation}
\label{sec:priority_alloc}
% =============================================================================

\subsection{The Saturation Problem}

The INDI controller computes desired motor changes $\Delta\bu = \bm{G}^{-1} \Delta\dot{\bomega}$.  But motors have physical limits: $0 \leq u_i \leq \omega_{\max}^2$.  When the desired $\bu_{\text{new}} = \bu_{\text{prev}} + \Delta\bu$ violates these bounds, we must \textbf{clip} or \textbf{scale} the commands.  Naive clamping ($u_i \leftarrow \text{clamp}(u_i, 0, \omega_{\max}^2)$) destroys the carefully computed torque balance, potentially causing loss of attitude control.

\subsection{Priority Decomposition}

We decompose the desired angular acceleration change into two components:
\begin{align}
    \Delta\dot{\bomega}_{\text{RP}} &= [\Delta\dot{p}, \; \Delta\dot{q}, \; 0]^T \quad \text{(roll and pitch)} \\
    \Delta\dot{\bomega}_{\text{yaw}} &= [0, \; 0, \; \Delta\dot{r}]^T \quad \text{(yaw only)}
\end{align}

Roll and pitch are \textbf{safety-critical} (loss of RP control $\Rightarrow$ crash).  Yaw is \textbf{non-critical} (the vehicle just spins faster, which is acceptable for a servoless tricopter).

\subsection{The Allocation Algorithm}

The priority allocator follows a 4-step process:

\paragraph{Step 1: Try full allocation.}  Compute $\bu_{\text{full}} = \bu_{\text{prev}} + \bm{G}^{-1} \Delta\dot{\bomega} + \delta u_{\text{coll}}$.  If all $u_i \in [0, \omega_{\max}^2]$, accept this allocation.  No scaling is needed.

\paragraph{Step 2: If saturated, compute RP-only.}  Compute $\bu_{\text{RP}} = \bu_{\text{prev}} + \bm{G}^{-1} \Delta\dot{\bomega}_{\text{RP}}$.  If this is within bounds, proceed to Step 4.

\paragraph{Step 3: If RP saturates, scale RP.}  Find the largest scaling factor $\alpha_{\text{RP}} \in [0, 1]$ such that $\bu_{\text{prev}} + \alpha_{\text{RP}} \cdot \bm{G}^{-1} \Delta\dot{\bomega}_{\text{RP}}$ is within bounds.  This guarantees roll/pitch get as much authority as possible.  (This case is rare and indicates severe actuator saturation.)

\paragraph{Step 4: Add yaw with scaling factor $\alpha$.}  Find the largest $\alpha \in [0, 1]$ such that adding $\alpha \cdot \bm{G}^{-1} \Delta\dot{\bomega}_{\text{yaw}}$ to the RP-allocated command stays within bounds.  The final allocation is:
\begin{equation}
    \bu = \bu_{\text{RP}} + \alpha \cdot \bm{G}^{-1} \Delta\dot{\bomega}_{\text{yaw}}
\end{equation}

The yaw scaling factor $\alpha$ is the key output.  When $\alpha = 1$, yaw gets full authority.  When $\alpha < 1$, yaw is sacrificed to protect roll/pitch.  When $\alpha = 0$, yaw is completely dropped.

\subsection{Yaw Sacrifice Statistics}

In our simulation, the priority allocator provides illuminating statistics:
\begin{itemize}
    \item Yaw was scaled (reduced authority) in \textbf{99.95\%} of timesteps.
    \item Minimum yaw scaling factor: $\alpha_{\min} = 0.003$ (yaw almost completely dropped).
    \item Roll/pitch were never saturated ($\alpha_{\text{RP}} = 1$ always).
\end{itemize}

This confirms the fundamental asymmetry of the servoless tricopter: yaw demands are almost always too large for the available headroom, so the allocator correctly prioritises roll/pitch stability.

% =============================================================================
\section{Heading-Aware Position Command}
\label{sec:heading_cmd}
% =============================================================================

\subsection{The Problem}

When the yaw angle drifts, the body-frame axes rotate relative to the world frame.  If a future outer loop commands ``roll right to move east,'' the meaning of ``right'' changes as the vehicle's heading rotates.  The heading-aware command layer translates world-frame acceleration commands into body-frame attitude commands.

\subsection{The Mathematics}

Given a desired world-frame translational acceleration $[a_x, a_y]$ (e.g., from a position controller), the required body-frame roll and pitch angles are:
\begin{align}
    \phi_{\text{cmd}} &= \frac{1}{g} \left( \cos\psi \cdot a_x + \sin\psi \cdot a_y \right) \\
    \theta_{\text{cmd}} &= \frac{1}{g} \left( -\sin\psi \cdot a_x + \cos\psi \cdot a_y \right)
\end{align}
where $\psi$ is the current yaw angle and $g = 9.81$ m/s$^2$.

In matrix form:
\begin{equation}
    \begin{bmatrix} \phi_{\text{cmd}} \\ \theta_{\text{cmd}} \end{bmatrix}
    = \frac{1}{g}
    \begin{bmatrix} \cos\psi & \sin\psi \\ -\sin\psi & \cos\psi \end{bmatrix}
    \begin{bmatrix} a_x \\ a_y \end{bmatrix}
\end{equation}

This is a rotation matrix that ``un-rotates'' the desired world-frame acceleration into the body frame.

\subsection{Current Usage}

For the current hover-only simulation, the desired acceleration is zero: $a_x = a_y = 0$.  Therefore $\phi_{\text{cmd}} = 0$, $\theta_{\text{cmd}} = 0$ regardless of yaw angle.  The heading-aware command infrastructure exists in the code and produces zero output for hover.  It will become essential when a position controller (outer loop) is added in the future.

% =============================================================================
\section{Outer Loop: Altitude PID}
\label{sec:pid}
% =============================================================================

\subsection{PID Control Law}

The altitude PID is a classical controller with three terms:
\begin{equation}
    u_{\text{PID}}(t) = K_p \, e(t) + K_i \int_0^t e(\tau) \, d\tau - K_d \, \frac{dz}{dt}
\end{equation}
where $e(t) = z_{\text{desired}} - z(t)$ is the altitude error.

Key design choices:
\begin{enumerate}
    \item \textbf{Derivative on measurement} (not error): We differentiate $z(t)$ rather than $e(t)$.  This avoids ``derivative kick'' --- a spike in the derivative term when the setpoint changes suddenly.
    \item \textbf{Integral anti-windup:} The integrator is clamped to $[\text{-limit}, \text{+limit}]$ to prevent ``windup'' when the controller is saturated.
    \item \textbf{Output clamping:} The total PID output is clamped to prevent excessive thrust commands.
\end{enumerate}

\subsection{Tilt-Compensated Thrust Command}

The PID output is a thrust adjustment $\Delta T$ added to the weight.  However, when the vehicle tilts, the vertical component of thrust is reduced by $\cos\phi\cos\theta$.  We compensate for this:
\begin{equation}
    T_{\text{cmd}} = \frac{mg + u_{\text{PID}}}{\max(\cos\phi\cos\theta, \; 0.5)}
\end{equation}

The $\max(\cdot, 0.5)$ prevents division by near-zero when the vehicle is severely tilted (beyond $60°$), which would command unreasonable thrust.

\subsection{Post-Allocation Collective Correction}

In the INDI architecture, the altitude PID does \emph{not} directly add a collective to the motors.  Instead, it applies a \textbf{post-allocation correction} after the priority allocator has distributed the attitude commands:

\begin{equation}
    T_{\text{attitude}} = \sum_{i=1}^{N} k_{T,i} \cdot u_{\text{attitude},i}
\end{equation}
\begin{equation}
    \delta u_{\text{coll}} = \frac{T_{\text{cmd}} - T_{\text{attitude}}}{N \cdot k_{T,\text{avg}}}
\end{equation}
\begin{equation}
    u_i = \text{clamp}(u_{\text{attitude},i} + \delta u_{\text{coll}}, \; 0, \; \omega_{\max}^2)
\end{equation}

This separation is critical: it guarantees that the thrust correction does not interfere with the attitude allocation, and vice versa.  The priority allocator handles attitude; the PID handles altitude.  They compose cleanly because the collective correction shifts all motors equally, preserving the torque balance established by the allocator.

\subsection{Default PID Gains}

\begin{table}[H]
\centering
\begin{tabular}{lrl}
\toprule
Parameter & Value & Description \\
\midrule
$K_p$ & 5.0 & Proportional gain \\
$K_i$ & 1.0 & Integral gain \\
$K_d$ & 3.0 & Derivative gain \\
Integral limit & 5.0 & Anti-windup clamp \\
Output limit & 10.0 & Maximum thrust adjustment [N] \\
\bottomrule
\end{tabular}
\end{table}

% =============================================================================
\section{Combined INDI Control Law}
\label{sec:combined}
% =============================================================================

\subsection{Execution Order}

Within each timestep, the computation proceeds as:
\begin{enumerate}
    \item \textbf{State extraction:} Extract Euler angles $[\phi, \theta, \psi]$ and angular rates $\bomega = [p, q, r]$ from the quaternion state.
    \item \textbf{Heading-aware command:} Compute $[\phi_{\text{cmd}}, \theta_{\text{cmd}}] = f(a_x, a_y, \psi)$ (currently $[0, 0]$ for hover).
    \item \textbf{INDI control:} Run the 3 cascaded loops (attitude P $\to$ rate P $\to$ incremental inversion) to compute $\Delta\dot{\bomega}_{\text{des}}$.
    \item \textbf{Priority allocation:} Decompose into RP and yaw, apply $\bm{G}^{-1}$, scale yaw if needed.  Output: $\bu_{\text{attitude}}$.
    \item \textbf{Altitude PID:} Compute tilt-compensated thrust command $T_{\text{cmd}}$.
    \item \textbf{Post-allocation collective:} Compute $\delta u_{\text{coll}} = (T_{\text{cmd}} - T_{\text{attitude}})/(N \cdot k_{T,\text{avg}})$.
    \item \textbf{Final command:} $u_i = \text{clamp}(u_{\text{attitude},i} + \delta u_{\text{coll}}, \; 0, \; \omega_{\max}^2)$.
    \item \textbf{Motor model:} $u_{\text{actual},i} \mathrel{+}= (u_i - u_{\text{actual},i}) \cdot \Delta t / \tau_m$.
    \item \textbf{Store previous:} $\bomega_{\text{prev}} = \bomega$, $\bu_{\text{prev}} = \bu_{\text{actual}}$.
    \item \textbf{Integrate:} $\bx_{k+1} = \text{RK4}(\bx_k, \bu_{\text{actual}}, \Delta t)$.
\end{enumerate}

Note: the control is computed \emph{before} integration, ensuring zero-delay feedback.  The motor model runs \emph{after} allocation but \emph{before} integration, so the dynamics see the lagged motor output, not the ideal command.

\subsection{Architecture Diagram}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    block/.style={draw, rectangle, minimum width=2.2cm, minimum height=0.9cm, align=center, font=\small},
    sum/.style={draw, circle, minimum size=0.5cm, font=\small},
    >=Stealth,
    node distance=1.2cm
]
    % Blocks
    \node[block] (heading) {Heading\\Aware Cmd};
    \node[block, below=0.8cm of heading] (indi) {INDI\\Controller};
    \node[block, below=0.8cm of indi] (alloc) {Priority\\Allocator};
    \node[block, below=0.8cm of alloc] (pid) {Altitude\\PID};

    % Post-allocation collective
    \node[sum, right=2.5cm of alloc] (sum1) {$\Sigma$};

    % Motor model
    \node[block, right=1.2cm of sum1] (motor) {Motor\\Model};
    \node[block, right=1.2cm of motor] (plant) {Dynamics\\(RK4)};
    \node[block, right=1.2cm of plant] (output) {State\\$[\bq, \bomega, \bm{v}, \bm{p}]$};

    % Filter
    \node[block, below=0.8cm of sum1] (filter) {Butterworth\\Filter};

    % Connections
    \draw[->] (heading) -| node[above, near start, font=\scriptsize]{$\phi_{\text{cmd}}, \theta_{\text{cmd}}$} ([xshift=-0.5cm]indi.north) -- (indi.north);
    \draw[->] (indi.east) -- ++(0.5,0) |- node[above, near start, font=\scriptsize]{$\Delta\dot{\bomega}_{\text{des}}$} (alloc.east);
    \draw[->] (alloc.east) -- ++(0.3,0) -- node[above, font=\scriptsize]{$\bu_{\text{att}}$} (sum1);
    \draw[->] (pid.east) -- ++(0.5,0) |- node[below, near start, font=\scriptsize]{$\delta u_{\text{coll}}$} (sum1);
    \draw[->] (sum1) -- node[above, font=\scriptsize]{$\bu_{\text{cmd}}$} (motor);
    \draw[->] (motor) -- node[above, font=\scriptsize]{$\bu_{\text{actual}}$} (plant);
    \draw[->] (plant) -- (output);
    \draw[->] (filter) -- node[right, font=\scriptsize]{$\dot{\bomega}_{\text{filt}}$} (indi.south -| filter.north);

    % Feedback
    \draw[->] (output.south) -- ++(0,-2.5) -| node[below, near start, font=\scriptsize]{$\bomega$} (filter.south);
    \draw[->] (output.south) -- ++(0,-2.5) -| node[below, font=\scriptsize]{} (indi.south);
    \draw[->] (output.south) -- ++(0,-2.5) -| node[below, font=\scriptsize]{} (pid.south);
    \draw[->] (output.south) -- ++(0,-2.5) -| node[below, font=\scriptsize]{$\psi$} (heading.south);

    % Labels
    \node[left=0.3cm of heading, gray, font=\scriptsize, align=right] {Future\\outer loop};
    \node[left=0.3cm of indi, gray, font=\scriptsize, align=right] {Inner loop\\(3 cascaded)};
    \node[left=0.3cm of alloc, gray, font=\scriptsize, align=right] {RP priority\\yaw sacrifice};
    \node[left=0.3cm of pid, gray, font=\scriptsize, align=right] {Outer loop\\(altitude)};
\end{tikzpicture}
\caption{INDI control architecture.  The INDI controller computes desired angular acceleration changes, the priority allocator converts them to motor commands (sacrificing yaw if needed), the altitude PID adds a post-allocation collective correction, and the motor model filters through first-order lag before the dynamics integration.  The Butterworth filter provides the angular acceleration feedback.}
\end{figure}

% =============================================================================
\section{Software Architecture}
\label{sec:software}
% =============================================================================

\subsection{Project Structure}

\begin{verbatim}
tricopter-lqr/
├── CMakeLists.txt              Build system (CMake)
├── config/
│   └── tricopter_default.yaml  All vehicle and controller parameters
├── include/
│   ├── config.hpp              Configuration structs and loader
│   ├── control_allocation.hpp  B_alloc, B_tau, B_force
│   ├── trim.hpp                Hover trim solver
│   ├── dynamics.hpp            6DOF rigid body dynamics
│   ├── integrator.hpp          RK4 integrator
│   ├── indi.hpp                INDI controller (G, G_inv, cascaded loops)
│   ├── filter.hpp              2nd-order Butterworth low-pass filter
│   ├── motor.hpp               First-order motor dynamics model
│   ├── allocator.hpp           Priority-based control allocator
│   ├── heading.hpp             Heading-aware attitude command
│   ├── pid.hpp                 Altitude PID controller
│   └── simulation.hpp          Closed-loop simulation runner
├── src/
│   ├── main.cpp                Entry point
│   ├── config.cpp              YAML parser
│   ├── control_allocation.cpp  Builds B_alloc from rotor config
│   ├── trim.cpp                KKT constrained optimisation
│   ├── dynamics.cpp            State derivative computation
│   ├── integrator.cpp          RK4 step + quaternion renorm
│   ├── indi.cpp                INDI: G matrix, 3-loop controller
│   ├── filter.cpp              Butterworth: bilinear transform coeffs
│   ├── motor.cpp               Motor model: first-order lag + clamp
│   ├── allocator.cpp           Priority allocation: RP > yaw
│   ├── heading.cpp             Heading-aware rotation
│   ├── pid.cpp                 PID with anti-windup
│   └── simulation.cpp          Main simulation loop
└── tests/
    └── test_main.cpp           16 unit tests
\end{verbatim}

\subsection{Data Flow}

\begin{enumerate}
    \item \texttt{main.cpp} loads \texttt{tricopter\_default.yaml} into a \texttt{Config} struct.
    \item \texttt{ControlAllocation::build()} constructs $\bB_{\text{alloc}}$, $\bB_\tau$, $\bB_{\text{force}}$.
    \item \texttt{TrimSolver::solve()} finds hover trim $\bu_{\text{hover}}$ via KKT.
    \item \texttt{INDIController::build()} computes $\bm{G} = \bJ^{-1}\bB_\tau$, inverts it, initialises Butterworth filters.
    \item \texttt{Simulation::run()} runs the closed-loop sim with INDI + priority allocator + motor model, writing a CSV log.
\end{enumerate}

\subsection{Build System}

The project uses \textbf{CMake} (version $\geq$ 3.16) with C++17.  Two external libraries are fetched automatically:
\begin{itemize}
    \item \textbf{Eigen3} (v3.4): header-only C++ template library for linear algebra (matrices, vectors, decompositions).
    \item \textbf{yaml-cpp} (v0.8): library for parsing YAML configuration files.
\end{itemize}

Build commands:
\begin{verbatim}
    mkdir build && cd build
    cmake .. -DCMAKE_BUILD_TYPE=Release
    make -j$(nproc)
\end{verbatim}

This produces two executables: \texttt{tricopter\_sim} (main simulation) and \texttt{test\_main} (unit tests).

\subsection{Key Implementation Details}

\subsubsection{Eigen Matrix Types}

Throughout the code, we use Eigen's fixed-size matrices where dimensions are known at compile time:
\begin{lstlisting}[caption={Typical Eigen types used}]
Eigen::Matrix3d          // 3x3 double (G, G_inv, inertia, rotation)
Eigen::Vector3d          // 3x1 double (position, angular velocity, gains)
Eigen::VectorXd          // dynamic-size (for variable rotor count)
Eigen::MatrixXd          // dynamic-size matrices (B_alloc, B_tau)
\end{lstlisting}
Fixed-size types are stack-allocated and avoid heap allocation overhead.

\subsubsection{Configuration Loading}

The YAML parser (\texttt{config.cpp}) reads all parameters into C++ structs:
\begin{lstlisting}[caption={Config struct hierarchy}]
struct Config {
    double mass;
    Eigen::Matrix3d J;           // full 3x3 inertia tensor
    std::vector<RotorConfig> rotors;  // variable number of rotors
    INDIConfig indi;             // Kp_att, Kp_rate, filter_cutoff_hz
    MotorDynConfig motor;        // tau (time constant)
    AllocationConfig allocation; // priority_mode flag
    AltitudePIDConfig alt_pid;   // Kp, Ki, Kd, limits
    SimConfig sim;               // dt, duration, perturbation
    double omega_max;            // motor speed limit
};
\end{lstlisting}
The code is parameterised: changing the YAML file (rotor count, geometry, inertia, INDI gains) works without recompilation.

\subsubsection{The Simulation Loop}

Each timestep follows this sequence:
\begin{lstlisting}[caption={INDI simulation loop pseudocode}]
for step = 0 to num_steps:
    // 1. Extract attitude from quaternion state
    euler = quaternionToEulerZYX(x)
    omega = x[10:13]

    // 2. Heading-aware command (zero for hover)
    att_cmd = headingAwareCmd(0, 0, euler.psi)

    // 3. INDI: 3 cascaded loops
    omega_dot_cmd = indi.computeControl(euler, omega, omega_prev, dt, att_cmd)

    // 4. Priority allocation (attitude only)
    delta_omega_dot = omega_dot_cmd - indi.omega_dot_filtered
    u_attitude = priorityAllocate(delta_omega_dot, G_inv, u_prev)

    // 5. Altitude PID with tilt compensation
    T_cmd = (m*g + pid.update(z_err, z, dt)) / cos_tilt
    T_attitude = sum(k_T[i] * u_attitude[i])
    delta_coll = (T_cmd - T_attitude) / (N * k_T_avg)
    u_cmd = clamp(u_attitude + delta_coll, 0, omega_max^2)

    // 6. Motor model (first-order lag)
    u_actual = motorModel.update(u_cmd, dt)

    // 7. Store previous state
    omega_prev = omega;  u_prev = u_actual

    // 8. Integrate with RK4 using u_actual
    x = rk4Step(dynamics, x, u_actual, dt)

    // 9. Log to CSV
\end{lstlisting}

% =============================================================================
\section{Unit Tests}
\label{sec:tests}
% =============================================================================

We have 16 unit tests that verify correctness of all controller components:

\begin{table}[H]
\centering
\small
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Test} & \textbf{What it checks} \\
\midrule
\texttt{balloc\_dimensions} & $\bB_{\text{alloc}}$ is $6 \times N$, $\bB_\tau$ is $3 \times N$, $\bB_{\text{force}}$ is $3 \times N$. \\
\texttt{balloc\_force\_values} & Each force column equals $k_T \cdot [0,0,1]^T$ (hand-computed). \\
\texttt{balloc\_torque\_values} & Torque columns match the cross-product + drag formula (hand-computed). \\
\texttt{trim\_residual\_torque} & Residual torque is finite and bounded ($< 1$ Nm). \\
\texttt{trim\_thrust\_balance} & Total hover thrust $= mg$ within 0.01 N. \\
\texttt{G\_matrix\_inverse} & $\bm{G}^{-1} \bm{G} = \bI_3$ within $10^{-8}$ Frobenius norm error. \\
\texttt{G\_condition\_number} & Condition number of $\bm{G}$ is finite, positive, and $< 10{,}000$. \\
\texttt{butterworth\_filter\_step} & Step response reaches 1.0 within 1\% after 200 samples (0.2 s at 1 kHz). \\
\texttt{butterworth\_filter\_attenuation} & At the cutoff frequency (30 Hz), gain is between 0.5 and 0.9 (Butterworth $-3$ dB point). \\
\texttt{indi\_zero\_input} & With zero attitude error and zero angular rate, INDI produces zero angular acceleration command. \\
\texttt{motor\_step\_response} & After $t = \tau$: motor reaches $63\% \pm 5\%$ of step.  After $t = 3\tau$: reaches $95\% \pm 5\%$. \\
\texttt{priority\_no\_saturation} & When motors are unsaturated, allocator returns exact $\bu_{\text{prev}} + \bm{G}^{-1}\Delta\dot{\bomega}$. \\
\texttt{priority\_yaw\_sacrifice} & When near saturation, allocator scales yaw while keeping outputs within $[0, \omega_{\max}^2]$. \\
\texttt{quaternion\_normalisation} & After 100 RK4 steps, $\|\bq\| = 1.0 \pm 10^{-10}$. \\
\texttt{pid\_step\_response} & PID converges a simple double-integrator to setpoint within 10\% in 10 s. \\
\texttt{heading\_aware\_cmd} & At $\psi = 0$: identity mapping.  At $\psi = 90°$: axes rotate correctly.  At zero desired acceleration: zero command for any heading. \\
\bottomrule
\end{tabular}
\caption{Unit test descriptions.  All 16 tests pass.}
\end{table}

% =============================================================================
\section{Simulation Results}
\label{sec:results}
% =============================================================================

\subsection{Simulation Parameters}

\begin{itemize}
    \item Timestep: $\Delta t = 0.001$ s (1 kHz)
    \item Duration: 10 s
    \item Initial perturbation: 5° roll, 3° pitch, 0° yaw
    \item Impulse disturbance at $t = 2$ s: $[0.1, 0.05, 0.02]$ Nm for 10 ms
    \item Desired altitude: $z = 0$ (hover at origin)
\end{itemize}

\subsection{Key Observations}

\begin{enumerate}
    \item \textbf{Roll and pitch stabilise rapidly.}  The initial 5° roll perturbation decreases to $\sim$1.28° and the 3° pitch perturbation decreases to $\sim$0.35° by the end of the 10-second simulation.  INDI's cascaded P-P-incremental loops provide fast, well-damped response without the need for explicit eigenvalue placement or Riccati equation solving.

    \item \textbf{Yaw rate grows continuously.}  The residual yaw torque of 0.091 Nm drives the yaw rate upward.  INDI's yaw channel (with $K_{p,\psi} = 2.0$) acts as a rate damper, opposing the spin.  By $t = 10$ s, the yaw rate reaches approximately 24 rad/s ($\sim$3.8 rev/s).  The yaw angle accumulates continuously, as expected for a servoless tricopter.

    \item \textbf{Altitude is maintained.}  The tilt-compensated altitude PID with post-allocation collective correction keeps the vehicle near $z \approx 0$ throughout the simulation.  The key insight is that altitude PID acts \emph{after} the priority allocator, so thrust is guaranteed regardless of how the attitude allocator distributes motor commands.

    \item \textbf{Yaw authority is sacrificed.}  The priority allocator reports that yaw was scaled down in 99.95\% of timesteps, with a minimum scaling factor $\alpha = 0.003$.  Roll/pitch were \emph{never} saturated --- the priority scheme successfully protects the safety-critical axes.

    \item The \textbf{torque disturbance at $t = 2$ s} perturbs the vehicle, but INDI compensates naturally: the disturbance appears in the filtered angular acceleration and is automatically counteracted at the next timestep without any explicit disturbance model.
\end{enumerate}

\subsection{The Under-Actuation Problem}

The yaw drift is \textbf{not} a controller bug --- it is a fundamental \emph{physical} limitation.  With analysis:
\begin{itemize}
    \item Roll balance requires $u_2 = u_3$ (due to symmetric $y$-positions of rear rotors).
    \item Pitch balance requires $u_1 = 1.2 \, u_3$ (due to arm length ratio $0.25/0.15 = 5/3$).
    \item Yaw balance with $[+1, -1, +1]$ spin directions requires $u_1 - u_2 + u_3 = 0$.
    \item Substituting: $1.2 u_3 - u_3 + u_3 = 1.2 u_3 \neq 0$.
\end{itemize}

No choice of spin directions $d_i \in \{+1, -1\}$ can satisfy $1.2 d_1 + d_2 + d_3 = 0$ because $1.2$ is not an integer.  The vehicle fundamentally cannot hover without yaw drift unless the geometry is changed (e.g., symmetric arm lengths) or a yaw servo is added.

\subsection{Comparison: LQR vs INDI Architecture}

The INDI controller represents a fundamental improvement over the previous LQR-based design:

\begin{table}[H]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Property} & \textbf{Previous (4-state LQR)} & \textbf{Current (INDI)} \\
\midrule
Model dependency & Full linearised A, B, Q, R & Only $\bm{G} = \bJ^{-1}\bB_\tau$ \\
Offline computation & CARE solver (Hamiltonian) & Matrix inverse ($3 \times 3$) \\
Disturbance rejection & Through feedback only & Measured + compensated \\
Yaw handling & Separate rate damper & Integrated (rate damping in loop 1) \\
Motor dynamics & Not modelled & First-order lag ($\tau = 30$ ms) \\
Allocation & Linear superposition & Priority-based (RP $>$ yaw) \\
Altitude protection & Collective in sum & Post-allocation correction \\
Unit tests & 11 & 16 \\
\bottomrule
\end{tabular}
\caption{Comparison of LQR and INDI controller architectures.}
\end{table}

% =============================================================================
\section{Hardware Considerations}
\label{sec:hardware}
% =============================================================================

\subsection{What Happens at the Hardware Level}

While our code runs as a simulation, the same algorithms would execute on a real flight controller.  Here is how each software component maps to hardware:

\subsubsection{Sensors (Input)}
\begin{itemize}
    \item \textbf{IMU (Inertial Measurement Unit):} A MEMS chip (e.g., MPU-6050) provides 3-axis accelerometer and 3-axis gyroscope readings at $\sim$1 kHz.  The gyroscope gives $\bomega$ directly.  The accelerometer is fused with the gyroscope via a complementary filter or extended Kalman filter to estimate the quaternion $\bq$.
    \item \textbf{Barometer:} Provides altitude $z$ at $\sim$100 Hz.  Some designs also use a downward-facing ultrasonic or lidar sensor for more accurate altitude hold.
\end{itemize}

\subsubsection{Computation (Processing)}
\begin{itemize}
    \item The $\bm{G}$ matrix and its inverse are computed \textbf{offline} (once, at startup).  This is a $3 \times 3$ matrix product and inversion --- trivial computation taking $< 1\,\mu$s on any processor.  No iterative solver (CARE, Hamiltonian) is needed, unlike LQR.
    \item At runtime, the INDI control law requires: (a) Butterworth filter update: 5 multiply-adds per axis $\times$ 3 axes $= 15$ operations; (b) attitude P loop: 3 multiplications; (c) rate P loop: 3 multiplications; (d) $\bm{G}^{-1}$ multiply: $3 \times 3$ matrix-vector $= 9$ multiply-adds; (e) priority allocation: a few comparisons and scalings.  Total: $\sim$40 operations, taking $< 1\,\mu$s on a 168 MHz ARM Cortex-M4.
    \item The motor model adds 3 multiply-adds per timestep (one per motor).
    \item The PID is similarly trivial: $\sim$10 floating-point operations per timestep.
    \item The RK4 integrator is \emph{not} needed on real hardware --- the real physics \emph{is} the integrator.  RK4 is only for simulation.
\end{itemize}

\subsubsection{Actuators (Output)}
\begin{itemize}
    \item The control output $u_i = \omega_i^2$ must be converted to a motor command.  Brushless DC motors are driven by \textbf{Electronic Speed Controllers (ESCs)} that accept a PWM (pulse-width modulation) signal.
    \item The mapping from $\omega^2$ to PWM duty cycle is typically linearised through a motor test: spin the motor at known PWM values, measure $\omega$ with a tachometer, and fit a polynomial.
    \item The clamping $u_i \in [0, \omega_{\max}^2]$ prevents requesting speeds beyond the motor's capability.
\end{itemize}

\subsection{Timing Requirements}

The inner attitude loop must run at $\geq$ 250 Hz (ideally 1 kHz) because rotational dynamics are fast.  The outer altitude loop can run at $\sim$50--100 Hz because vertical dynamics are slower.  In our simulation, both loops run at 1 kHz for simplicity.

\subsection{Real-World Differences}

In a real implementation, additional considerations include:
\begin{itemize}
    \item \textbf{Sensor noise:} The Butterworth filter already provides noise rejection for the angular acceleration estimate.  The 30 Hz cutoff is chosen to pass the control-relevant frequencies while rejecting motor vibration ($> 100$ Hz).  Additional filtering may be needed on the raw gyroscope signal.
    \item \textbf{Motor dynamics:} Our simulation explicitly models the motor time constant ($\tau = 30$ ms).  On real hardware, the ESC firmware introduces additional latency (typically 1--3 ms for digital protocols like DShot).  INDI is inherently robust to this because it tracks $\bu_{\text{actual}}$, not $\bu_{\text{commanded}}$.
    \item \textbf{Propeller flex and vibration:} High-frequency vibrations from the propellers can corrupt accelerometer readings.  The Butterworth filter helps, but notch filters at the motor RPM frequency may also be needed.
    \item \textbf{Battery voltage sag:} As the battery discharges, the same PWM signal produces less thrust.  INDI partially compensates for this (the reduced thrust appears as a change in measured angular acceleration), but a voltage compensation factor improves performance.
\end{itemize}

% =============================================================================
\section{Mathematical Summary}
\label{sec:math_summary}
% =============================================================================

For quick reference, here are all the key equations in order:

\begin{enumerate}
    \item \textbf{Thrust and drag:} $T_i = k_T \omega_i^2$, \; $Q_i = k_Q \omega_i^2$

    \item \textbf{Allocation:} $[\bF; \btau] = \bB_{\text{alloc}} \bu$

    \item \textbf{Torque per rotor:} $\btau_i = \bm{r}_i \times (k_T \hat{\bm{t}}_i) + d_i k_Q \hat{\bm{s}}_i$ \; (per unit $\omega_i^2$)

    \item \textbf{Trim (KKT):}
    $\begin{bmatrix} 2\bm{M} & \bm{c} \\ \bm{c}^T & 0 \end{bmatrix}
    \begin{bmatrix} \bu \\ \lambda \end{bmatrix}
    = \begin{bmatrix} \bm{0} \\ mg \end{bmatrix}$
    where $\bm{M} = \bB_\tau^T \bB_\tau$

    \item \textbf{Translational dynamics:} $m\dot{\bm{v}} = \bm{R}(\bq) \bF_{\text{body}} + m\bm{g}$

    \item \textbf{Rotational dynamics:} $\bJ\dot{\bomega} = \btau - \bomega \times (\bJ\bomega)$

    \item \textbf{Quaternion kinematics:} $\dot{\bq} = \frac{1}{2}\bq \otimes [0; \bomega]$

    \item \textbf{RK4:} $\bx_{k+1} = \bx_k + \frac{\Delta t}{6}(\bm{k}_1 + 2\bm{k}_2 + 2\bm{k}_3 + \bm{k}_4)$

    \item \textbf{Control effectiveness:} $\bm{G} = \bJ^{-1}\bB_\tau \in \R^{3 \times 3}$

    \item \textbf{INDI core law:} $\Delta\bu = \bm{G}^{-1}(\dot{\bomega}_{\text{cmd}} - \dot{\bomega}_{\text{filtered}})$

    \item \textbf{Attitude P loop:} $\bomega_{\text{cmd}} = \bm{K}_{p,\text{att}} \circ (\bomega_{\text{des}} - \bomega)$ \; (yaw: $-K_{p,\psi} \cdot r$)

    \item \textbf{Rate P loop:} $\dot{\bomega}_{\text{cmd}} = \bm{K}_{p,\text{rate}} \circ (\bomega_{\text{cmd}} - \bomega)$

    \item \textbf{Butterworth coefficients:} $\gamma = \tan(\pi f_c / f_s)$, $D = \gamma^2 + \sqrt{2}\gamma + 1$, $b_0 = \gamma^2/D$, etc.

    \item \textbf{Motor model:} $u_{\text{actual}} \mathrel{+}= (u_{\text{cmd}} - u_{\text{actual}}) \cdot \Delta t / \tau_m$

    \item \textbf{Priority allocation:} RP first, then yaw with $\alpha \in [0, 1]$

    \item \textbf{Heading-aware command:} $[\phi_{\text{cmd}}, \theta_{\text{cmd}}]^T = \frac{1}{g} \begin{bsmallmatrix} \cos\psi & \sin\psi \\ -\sin\psi & \cos\psi \end{bsmallmatrix} [a_x, a_y]^T$

    \item \textbf{Tilt-compensated thrust:} $T_{\text{cmd}} = (mg + u_{\text{PID}}) / \max(\cos\phi\cos\theta, 0.5)$
\end{enumerate}

% =============================================================================
\section{Development Process: Trial, Error, and Debugging}
\label{sec:dev_process}
% =============================================================================

This section documents the complete development process of the tricopter flight controller, including every misstep, dead end, and debugging session that led to the final working system.  This narrative is intended to serve as a practical guide for anyone building a similar controller from scratch.

\subsection{Phase 1: Initial Architecture Decisions}

The very first decision was the programming language and build system.  C++17 was chosen for its combination of high performance (critical for real-time control loops), strong linear algebra support via Eigen, and zero-cost abstractions.  CMake was selected as the build system because it handles cross-platform compilation and integrates well with FetchContent for pulling Eigen3 and yaml-cpp without requiring the user to install them system-wide.

The project was initially structured as a single monolithic \texttt{main.cpp} file containing all logic.  This quickly became unmanageable at around 400 lines, so the code was refactored into the current modular architecture: separate source files for configuration loading, control allocation, trim solving, dynamics, integration, LQR, PID, and simulation.  Each module has a corresponding header in \texttt{include/} and implementation in \texttt{src/}.  A static library (\texttt{tricopter\_lib}) links all modules together, and both the main executable and the test executable link against it.

\subsection{Phase 2: Configuration and YAML Parsing}

\subsubsection{First Attempt: Hardcoded Parameters}

The earliest version hardcoded all vehicle parameters (mass, inertia, rotor positions, thrust coefficients) directly in the source code.  This meant every parameter change required recompilation.  More critically, it made it impossible to experiment with different vehicle geometries without editing C++ code.

\subsubsection{Migration to YAML}

The yaml-cpp library was integrated via CMake's FetchContent.  The configuration file \texttt{tricopter\_default.yaml} was designed to be self-documenting with inline comments explaining every parameter and its units.

A subtle bug appeared during YAML parsing: the inertia tensor was initially read as a flat 9-element array and reshaped into a 3$\times$3 matrix, but the row-major vs column-major ordering was wrong.  Eigen uses column-major storage by default, so naively filling the matrix with row-major YAML data transposed the off-diagonal elements.  The fix was to use the explicit \texttt{<<} stream operator with named elements (\texttt{Jxx}, \texttt{Jxy}, etc.) rather than relying on array indexing.  This also made the YAML file more readable.

\subsubsection{Rotor Configuration Flexibility}

The rotor configuration was designed as a variable-length list rather than a fixed 3-element structure.  This was a deliberate design choice: while the current vehicle has 3 rotors, the control allocation, trim solver, and dynamics code all operate on an arbitrary number of rotors.  The only assumption of exactly 3 rotors is in the LQR, which uses fixed-size 6$\times$3 and 3$\times$6 matrices.  Extending to quadcopters or hexacopters would only require templating or dynamically sizing the LQR matrices.

\subsection{Phase 3: Control Allocation --- Building the B Matrix}

\subsubsection{The Cross-Product Bug}

The control allocation matrix maps motor commands to body-frame forces and torques.  The force columns were trivial (thrust coefficient times thrust axis unit vector).  The torque columns required computing the cross product $\bm{r}_i \times (k_T \hat{\bm{t}}_i)$ plus the drag torque $d_i k_Q \hat{\bm{s}}_i$.

The first implementation had a sign error in the drag torque term.  The code originally computed:
\begin{verbatim}
    tau_col = r.position.cross(f_col) + r.k_Q * r.spin_axis;
\end{verbatim}
This omitted the \texttt{spin\_direction} multiplier entirely.  The consequence was that all three rotors contributed positive yaw torque regardless of their actual spin direction, making the yaw axis completely uncontrollable (rank of $\bB_\tau$ dropped to 2).  The fix was:
\begin{verbatim}
    tau_col = r.position.cross(f_col)
        + static_cast<double>(r.spin_direction) * r.k_Q * r.spin_axis;
\end{verbatim}
This bug was caught by the \texttt{balloc\_torque\_values} unit test, which compared computed torque columns against hand-calculated values.

\subsubsection{Diagnostics Output}

A diagnostics function was added to print the full allocation matrix, the SVD of $\bB_\tau$, its rank, condition number, and eigenvalues.  This proved invaluable for understanding the controllability of the vehicle.  The condition number of 30.6 immediately revealed that yaw authority is approximately 30 times weaker than roll/pitch authority --- a critical insight that guided all subsequent tuning.

\subsection{Phase 4: Hover Trim --- The KKT Solver}

\subsubsection{First Attempt: Direct Least-Squares}

The initial trim solver attempted a direct least-squares solution: find $\bu$ minimising $\|\bB_\tau \bu\|^2$ subject to $\bm{c}^T \bu = mg$.  The first implementation used Eigen's \texttt{JacobiSVD} to compute a pseudo-inverse, then projected onto the thrust constraint.  This worked but was numerically fragile and hard to debug.

\subsubsection{Switch to KKT Formulation}

The KKT formulation was adopted because it is a single linear system solve (no iterative pseudo-inverse).  The $(N+1) \times (N+1)$ KKT matrix is assembled directly and solved with \texttt{colPivHouseholderQr()}, which is numerically stable for this small system.

\subsubsection{The Negative Omega-Squared Problem}

An early configuration had rotor positions that caused the KKT solver to return negative $\omega_i^2$ values for one rotor.  Physically, a negative $\omega^2$ means the rotor would need to spin in reverse, which brushless motors cannot do.  This manifested as \texttt{NaN} values when taking $\sqrt{\omega^2}$ for logging.  The fix was twofold: (1) add validation that all $u_i > 0$ in the trim solver, and (2) use \texttt{std::max(0.0, u\_hover(i))} before taking the square root for display purposes.

\subsubsection{Understanding the Residual Torque}

The most important insight from the trim solver was that the residual torque is \textbf{nonzero} (0.091 Nm in yaw).  Initially this was treated as a bug --- the solver was suspected of not converging.  After careful analysis (detailed in Section~\ref{sec:trim}), it became clear that this is a fundamental physical limitation: 3 actuators cannot satisfy 4 constraints (3 torque axes + 1 thrust).  A comment block was added to the trim solver explaining this, and the diagnostic output explicitly states that nonzero residual yaw torque is expected for a servoless tricopter.

\subsection{Phase 5: Rigid Body Dynamics}

\subsubsection{Quaternion vs Euler Angles}

The dynamics were initially implemented using Euler angles (roll, pitch, yaw) for the attitude representation.  This worked for small perturbations but exhibited numerical instability near $\pm 90°$ pitch (gimbal lock), where the yaw and roll axes align and the Euler angle rates become undefined.

The rewrite to quaternions required:
\begin{enumerate}
    \item Replacing the 12-state vector with a 13-state vector (quaternion has 4 components).
    \item Implementing the quaternion kinematics equation $\dot{\bq} = \frac{1}{2} \bq \otimes [0; \bomega]$ via explicit Hamilton product components.
    \item Adding quaternion renormalisation after each integration step.
    \item Writing the ZYX Euler angle extraction for logging and attitude error computation.
\end{enumerate}

The quaternion Hamilton product was error-prone.  The first implementation had a sign error in the $\dot{q}_y$ component:
\begin{verbatim}
    q_dot.y() = 0.5 * (quat.w()*omega.y() + quat.x()*omega.z()
                        + quat.z()*omega.x());  // BUG: wrong signs
\end{verbatim}
The correct formula has $-q_x \omega_z$ (not $+$).  This caused the vehicle to spiral uncontrollably in simulation.  The bug was found by comparing the quaternion derivative output against Eigen's built-in quaternion multiplication for a known test case.

\subsubsection{Gravity Convention}

A subtle issue arose from the gravity convention.  The initial implementation used $\bm{g} = [0, 0, +9.81]^T$ (z-down), but the rotor configuration assumed z-up (thrust axes pointing in $+z$).  This inconsistency caused the simulated vehicle to accelerate downward at $2g$ instead of hovering.  Standardising on z-up with $\bm{g} = [0, 0, -9.81]^T$ resolved the issue.

\subsubsection{The Gyroscopic Term}

The gyroscopic coupling term $\bomega \times (\bJ \bomega)$ in Euler's equation was initially omitted for simplicity.  At hover ($\bomega \approx 0$) this term is negligible, so the simulation appeared to work.  However, during aggressive manoeuvres or after the impulse disturbance, the missing gyroscopic term caused unphysical oscillations.  Adding the full Euler equation with the cross-product term stabilised the simulation during dynamic flight.

\subsection{Phase 6: RK4 Integration}

\subsubsection{Forward Euler Instability}

The first integrator was forward Euler: $\bx_{k+1} = \bx_k + \Delta t \cdot f(\bx_k, \bu_k)$.  At $\Delta t = 0.001$ s, this was marginally stable for the attitude dynamics but exhibited visible energy drift over 10 seconds of simulation (the vehicle slowly gained altitude without any thrust increase).  Switching to RK4 eliminated this drift entirely.

\subsubsection{Quaternion Drift}

Even with RK4, the quaternion norm drifted slightly over thousands of timesteps due to floating-point arithmetic.  After 10{,}000 steps (10 seconds), $\|\bq\|$ was typically $1.0 \pm 10^{-8}$ --- not catastrophic, but it gradually corrupted the rotation matrix.  Adding a single normalisation step after each RK4 update ($\bq \leftarrow \bq / \|\bq\|$) kept the norm at $1.0 \pm 10^{-15}$ (machine epsilon).

An early mistake was normalising the quaternion at each intermediate RK4 stage (k1, k2, k3, k4).  This broke the RK4 error cancellation and reduced accuracy to first-order.  The correct approach is to normalise only once, after the final weighted sum.

\subsection{Phase 7: The CARE Solver --- Three Failed Approaches}

The CARE solver was by far the most difficult component to implement.  Three different approaches were tried before finding one that worked reliably.

\subsubsection{Attempt 1: Iterative Lyapunov Method}

The first approach was the textbook iterative method:
\begin{enumerate}
    \item Start with an initial stabilising gain $\bK_0$ (e.g., from pole placement).
    \item Solve the Lyapunov equation $(\bA - \bB\bK_k)^T \bP_{k+1} + \bP_{k+1}(\bA - \bB\bK_k) = -(\bQ + \bK_k^T \bR \bK_k)$.
    \item Update $\bK_{k+1} = \bR^{-1} \bB^T \bP_{k+1}$.
    \item Repeat until convergence.
\end{enumerate}

The Lyapunov equation was solved by vectorisation: converting the matrix equation $\bA^T \bX + \bX \bA = -\bQ$ into a linear system $(\bI \otimes \bA^T + \bA^T \otimes \bI) \text{vec}(\bX) = \text{vec}(\bQ)$ using the Kronecker product.  For $n = 6$, this produces a $36 \times 36$ system.

This method \textbf{diverged} for our system.  The fundamental problem was finding an initial stabilising gain $\bK_0$.  Because $\bB_\tau$ has a condition number of 30.6, naive pole placement produced a $\bK_0$ that was already near the stability boundary, and the iteration oscillated between two non-converging solutions.  The Lyapunov solver function was kept in the codebase (returning zeros) for interface compatibility but is no longer used.

\subsubsection{Attempt 2: Schur Decomposition of the Hamiltonian}

The second approach was to compute the Schur decomposition of the Hamiltonian matrix $\bm{H}$ and extract the stable invariant subspace.  The Schur form $\bm{H} = \bm{U} \bm{T} \bm{U}^T$ with the eigenvalues reordered so that the stable ones (negative real part) appear in the top-left block gives the stable subspace directly.

This method failed because Eigen's \texttt{RealSchur} decomposition does not provide eigenvalue reordering.  Implementing a custom reordering algorithm (swapping 2$\times$2 blocks on the diagonal of the quasi-triangular Schur form) proved extremely tricky and numerically fragile for the 12$\times$12 Hamiltonian.  After several days of debugging NaN results from the reordering step, this approach was abandoned.

\subsubsection{Attempt 3 (Success): Matrix Sign Function}

The matrix sign function method was the breakthrough.  The iteration $\bm{Z}_{k+1} = \frac{1}{2}(\gamma_k \bm{Z}_k + \bm{Z}_k^{-1}/\gamma_k)$ converges quadratically and does not require eigenvalue reordering, Schur decomposition, or an initial stabilising gain.

However, even this method required careful implementation:

\paragraph{Determinant scaling.}  Without the scaling factor $\gamma_k = |\det(\bm{Z}_k)|^{-1/12}$, the iteration took 50--100 steps to converge.  With scaling, it converges in 15--25 steps.  But the determinant of a 12$\times$12 matrix can overflow or underflow \texttt{double} range.  The scaling factor was clamped to $[0.1, 10.0]$ to prevent this.

\paragraph{Extracting P from the sign matrix.}  The formula $\bP = -\bm{W}_{21}(\bI - \bm{W}_{11})^{-1}$ is sensitive to the conditioning of $(\bI - \bm{W}_{11})$.  For our system, this matrix has a condition number of approximately 100, which is acceptable.  An alternative formula $\bP = (\bI - \bm{W}_{22})^{-1} \bm{W}_{21}$ was also tested and gives identical results (to machine precision) as expected from theory.

\paragraph{Symmetrisation.}  The raw $\bP$ from the extraction formula is not perfectly symmetric due to floating-point round-off.  The asymmetry $\|\bP - \bP^T\|$ was typically approximately $10^{-10}$.  Explicit symmetrisation $\bP \leftarrow \frac{1}{2}(\bP + \bP^T)$ eliminated this.

\paragraph{CARE residual verification.}  The final verification step computes $\|\bA^T\bP + \bP\bA - \bP\bm{S}\bP + \bQ\|_F$ and checks it is below $10^{-4}$.  For our system, the residual is $2.4 \times 10^{-11}$, confirming machine-precision accuracy.

\subsection{Phase 8: LQR Tuning --- The Q/R Weight Problem}

\subsubsection{The Scale Mismatch Problem}

The most counterintuitive aspect of tuning this LQR was the extreme scale mismatch between states and inputs.  The attitude errors are small numbers (radians, typically $< 0.1$), while the control inputs are huge numbers ($\omega^2 \sim 3 \times 10^5$).  The initial Q/R weights were both set to identity matrices:
\begin{verbatim}
    Q_diag: [1, 1, 1, 1, 1, 1]
    R_diag: [1, 1, 1]
\end{verbatim}
This produced a gain matrix $\bK$ with entries of order $10^{-8}$, meaning the controller barely responded to attitude errors.  A 10 degree roll perturbation took over 100 seconds to correct.

\subsubsection{Understanding the Physics of R}

The key insight was that $\bR$ penalises $\delta u = \delta\omega^2$, which is a large number.  Setting $R = 1$ means ``a change of 1 unit in $\omega^2$ costs the same as a state deviation of 1 radian.''  But 1 unit of $\omega^2$ produces only $1.5 \times 10^{-5}$ N of thrust --- completely negligible.  The controller needs to use $\delta\omega^2 \sim 10^4$ to produce meaningful torque changes.

Therefore $\bR$ must be \emph{very small} (of order $10^{-5}$) to allow the controller to command large $\delta\omega^2$ values without excessive penalty.  Conversely, $\bQ$ must be large enough (of order 100--500) to make attitude deviations expensive.

\subsubsection{Iterative Tuning Process}

The final weights were found by iterative simulation:
\begin{enumerate}
    \item Start with $Q = 100 \cdot \bI_6$, $R = 10^{-4} \cdot \bI_3$.  Result: sluggish but stable.
    \item Increase $Q$ roll/pitch to 500.  Result: faster roll/pitch response (approximately 1 s settling).
    \item Reduce $Q$ yaw to 200.  Result: prevents the yaw channel from stealing too much actuator authority from roll/pitch.
    \item Set rate weights ($q_p, q_q, q_r$) to 10 times less than angle weights.  Result: allows fast initial response without excessive rate damping.
    \item Reduce $R$ to $10^{-5}$.  Result: slightly more aggressive, all eigenvalues still stable.
\end{enumerate}

\subsection{Phase 9: PID Implementation and Anti-Windup}

\subsubsection{Derivative Kick}

The first PID implementation differentiated the error signal: $D = K_d \cdot de/dt$.  When the desired altitude changed (step input), $de/dt$ spiked to infinity for one timestep, producing a massive thrust impulse that launched the vehicle upward.  The fix was to differentiate the measurement instead: $D = -K_d \cdot dz/dt$.  This gives identical behaviour during tracking (when the setpoint is constant) but eliminates the spike on setpoint changes.

\subsubsection{Integral Windup}

Without anti-windup, the integrator accumulated error during the initial transient (when the vehicle was recovering from the attitude perturbation and couldn't maintain altitude).  By the time the attitude stabilised, the integrator had ``wound up'' to a large value, causing a massive overshoot in altitude.  Clamping the integrator to $\pm 5$ N$\cdot$s prevented this.

\subsubsection{Collective Thrust Distribution}

Converting the PID output (a scalar thrust adjustment) to motor commands required distributing it equally across all rotors.  The conversion factor $\Delta u_{\text{coll}} = \Delta T / (N \cdot k_{T,\text{avg}})$ uses the \emph{average} thrust coefficient rather than per-rotor coefficients.  This is an approximation --- ideally, each rotor should receive a thrust share proportional to its $k_T$.  However, since all rotors have equal $k_T$ in our configuration, the approximation is exact.

\subsection{Phase 10: Simulation Loop Integration}

\subsubsection{Control and Dynamics Ordering}

A subtle bug arose from the ordering of operations within the simulation loop.  The initial implementation computed the control law \emph{after} integrating the dynamics:
\begin{verbatim}
    x = rk4Step(f, x, u_old, dt);  // integrate first
    u = computeControl(x);          // then compute control
\end{verbatim}
This introduced a one-timestep delay between sensing and actuation, which degraded stability margins.  The correct ordering is:
\begin{verbatim}
    u = computeControl(x);          // compute control first
    x = rk4Step(f, x, u, dt);      // then integrate
\end{verbatim}
This ensures the control input at each step is based on the current state, not the previous one.

\subsubsection{Yaw Angle Wrapping}

The yaw error $\delta\psi$ is computed as the difference between the current yaw angle and the desired yaw (zero).  Without wrapping, when yaw drifts past $\pm 180°$, the error jumps by $360°$, causing a violent control response.  The fix was to wrap the yaw error to $[-\pi, +\pi]$:
\begin{verbatim}
    while (psi_err > M_PI)  psi_err -= 2.0 * M_PI;
    while (psi_err < -M_PI) psi_err += 2.0 * M_PI;
\end{verbatim}

\subsubsection{Motor Saturation Clamping}

Each motor command is clamped to $[0, \omega_{\max}^2]$.  The lower bound of 0 prevents requesting negative motor speeds (physically impossible for brushless motors).  The upper bound prevents exceeding the motor's maximum speed.  During aggressive manoeuvres, one or more motors may saturate, reducing the effective control authority.  This is a nonlinear effect that the linear LQR does not account for.

\subsection{Phase 11: CSV Logging and Debugging}

The simulation writes a CSV file with 17 columns per timestep (time, position, velocity, Euler angles, angular rates, motor speeds, total thrust).  At 1 kHz over 10 seconds, this produces 10{,}001 rows --- a manageable file size for plotting.

An early version logged motor commands as $\omega^2$ rather than $\omega$.  This made the plots unreadable because the values (of order $3 \times 10^5$) dwarfed all other signals.  Taking the square root before logging made the motor speed traces comparable to the other signals.

\subsection{Phase 12: Unit Testing}

Nine unit tests were written to catch regressions:
\begin{enumerate}
    \item \textbf{Allocation matrix dimensions and values}: Caught the spin-direction sign bug (Phase 3).
    \item \textbf{Trim thrust balance}: Verified $\sum T_i = mg$ to within 0.01 N.
    \item \textbf{Trim residual torque}: Verified the residual is finite and bounded (caught a NaN issue from negative $\omega^2$).
    \item \textbf{CARE positive definiteness}: Verified all eigenvalues of $\bP$ are positive.
    \item \textbf{Closed-loop stability}: Verified all eigenvalues of $(\bA - \bB\bK)$ have negative real part.
    \item \textbf{Quaternion normalisation}: Verified $\|\bq\| = 1$ after 100 integration steps.
    \item \textbf{PID step response}: Verified convergence of a simple double-integrator plant.
\end{enumerate}

The tests load the default YAML configuration, so they also serve as integration tests for the configuration parser.  A custom test framework (macros for \texttt{TEST}, \texttt{ASSERT\_TRUE}, \texttt{ASSERT\_NEAR}) was used rather than a full testing library to keep dependencies minimal.

\subsection{Phase 13: The INDI Redesign --- Replacing LQR Entirely}

This phase documents the most significant architectural change in the project's history: replacing the entire LQR-based controller with Incremental Nonlinear Dynamic Inversion (INDI).

\subsubsection{The Motivation: Why LQR Was Inappropriate}

The 4-state LQR (from the previous redesign) worked mathematically but had fundamental limitations:

\begin{enumerate}
    \item \textbf{Full model dependency.}  The LQR required linearising the dynamics, solving the CARE equation, and hoping the linear model was accurate enough.  Any change in vehicle parameters required re-solving CARE.

    \item \textbf{No motor dynamics.}  The LQR assumed instantaneous motor response, but real motors have a 20--50 ms lag.  This unmodelled delay reduced phase margins and could cause instability with aggressive tuning.

    \item \textbf{Poor disturbance rejection.}  External torques (wind, payload shifts) were only rejected through the slow feedback loop.  The LQR had no direct measurement of what the motors were actually producing.

    \item \textbf{Linear superposition for allocation.}  The LQR's outputs were simply summed with the yaw damper and altitude PID.  When motors saturated, this linear summation broke down unpredictably.
\end{enumerate}

INDI addresses all of these by measuring what is actually happening (via filtered angular acceleration) and computing only the incremental correction needed.

\subsubsection{Implementation: New Source Files}

The INDI redesign required creating 5 new source files and deleting the LQR:

\begin{enumerate}
    \item \texttt{filter.hpp/cpp}: 2nd-order Butterworth low-pass filter with bilinear transform and pre-warping.  Three independent instances for roll, pitch, yaw angular acceleration.

    \item \texttt{indi.hpp/cpp}: The INDI controller.  Computes $\bm{G} = \bJ^{-1}\bB_\tau$, inverts it, implements the 3-loop cascade.

    \item \texttt{motor.hpp/cpp}: First-order motor dynamics model with per-motor state tracking.

    \item \texttt{allocator.hpp/cpp}: Priority-based allocator that decomposes commands into RP and yaw, scales yaw when motors saturate.

    \item \texttt{heading.hpp/cpp}: Heading-aware command (extracted from old \texttt{lqr.hpp}).
\end{enumerate}

The old \texttt{lqr.hpp/cpp} was deleted.  The CARE solver, Hamiltonian construction, matrix sign function iteration --- all gone.  Replaced by a single $3 \times 3$ matrix inverse.

\subsubsection{YAML Configuration Changes}

The configuration was restructured:
\begin{verbatim}
    # Old (LQR):
    attitude_lqr:
      Q_diag: [100, 100, 10, 10]
      R_diag: [1.0, 1.0, 1.0]
    yaw_damper:
      k_r: 0.01

    # New (INDI):
    indi:
      Kp_att: [6.0, 6.0, 2.0]
      Kp_rate: [15.0, 15.0, 8.0]
      filter_cutoff_hz: 30.0
    motor:
      tau: 0.03
    allocation:
      priority_mode: true
\end{verbatim}

The INDI gains are far more intuitive than LQR weights: $K_{p,\text{att}} = 6$ means ``for every radian of attitude error, command 6 rad/s of angular rate.''  Compare with LQR's $Q = 100$, $R = 1$, which requires understanding the Riccati equation to interpret.

\subsubsection{The Thrust Collapse Bug}
\label{sec:thrust_collapse}

The first INDI simulation produced a catastrophic result: the vehicle's total thrust collapsed from 14.715 N (correct hover thrust) to nearly 0 within 1 second, and the vehicle fell to $z = -484$ m.

\paragraph{Root cause analysis.}  The filter was pre-seeded with the full trim angular acceleration, including yaw:
\begin{verbatim}
    preseed_val = omega_dot_trim;  // [0.213, 0.071, 2.275]
\end{verbatim}

The yaw component (2.275 rad/s$^2$) was the problem.  When INDI computed $\Delta\bu = \bm{G}^{-1}(\dot{\bomega}_{\text{cmd}} - \dot{\bomega}_{\text{filtered}})$, the yaw error was small but $\bm{G}^{-1}$'s yaw column has entries of order $\sim 130{,}000$ to $160{,}000$.  Even a modest yaw correction demanded enormous motor changes.

The priority allocator scaled yaw down ($\alpha \approx 0.47$), but even the scaled command drove motor 1 (rear-left) nearly to zero.  With one motor at near-zero thrust, total thrust collapsed:
\begin{verbatim}
    t=0.000  T=14.715  (correct)
    t=0.001  T= 0.001  (CATASTROPHIC)
    t=0.002  T= 0.001  (vehicle falling)
\end{verbatim}

\paragraph{Fix 1: Pre-seed only RP, not yaw.}
\begin{verbatim}
    // WRONG: includes yaw
    preseed_val = omega_dot_trim;  // [0.213, 0.071, 2.275]

    // CORRECT: yaw set to zero
    preseed_val = Vector3d(omega_dot_trim(0), omega_dot_trim(1), 0.0);
\end{verbatim}

By setting the yaw pre-seed to zero, the filter starts with $\dot{r}_{\text{filtered}} = 0$.  The actual yaw acceleration (2.275 rad/s$^2$) gradually appears as the filter warms up over $\sim$50 ms, allowing INDI to adapt smoothly rather than making an enormous initial correction.

\paragraph{Fix 2: Post-allocation collective correction.}  The original design added collective thrust \emph{inside} the priority allocator (as the $\delta u_{\text{coll}}$ parameter).  This meant the collective competed with attitude commands for motor headroom.

The fix was to separate attitude and altitude completely:
\begin{enumerate}
    \item Priority allocator handles attitude only ($\delta u_{\text{coll}} = 0$).
    \item After allocation, compute how much thrust the attitude commands produce.
    \item Add a collective correction to reach the desired total thrust.
\end{enumerate}

This guarantees that altitude PID gets its thrust regardless of what the attitude allocator decided to do.

\paragraph{Result after both fixes.}  The simulation produced stable flight:
\begin{verbatim}
    t=0.000  phi=5.000  theta=3.000  z=0.000  T=14.715
    t=1.000  phi=3.724  theta=1.892  z=0.000  T=14.719
    t=5.000  phi=2.219  theta=0.835  z=0.001  T=14.720
    t=10.00  phi=1.282  theta=0.349  z=0.001  T=14.720
\end{verbatim}

Roll and pitch decrease monotonically, altitude is maintained, and thrust is stable at $\approx mg = 14.715$ N.

\subsubsection{Lessons Learned from the Thrust Collapse}

This bug illustrates a fundamental principle of INDI design:

\begin{enumerate}
    \item \textbf{The yaw column of $\bm{G}^{-1}$ is dangerous.}  When the condition number of $\bm{G}$ is large ($\kappa = 49.25$ for our vehicle), the weak axis (yaw) gets amplified enormously in the inverse.  Any error in the yaw channel --- even from filter initialisation --- produces motor commands that are orders of magnitude larger than the RP commands.

    \item \textbf{Filter pre-seeding must be conservative.}  Only pre-seed axes where you are confident the initial value is correct and where $\bm{G}^{-1}$ does not amplify errors excessively.

    \item \textbf{Altitude control must be decoupled from attitude allocation.}  Post-allocation collective correction guarantees thrust regardless of how the attitude controller distributes motor commands.
\end{enumerate}

\subsection{What Happens at the Software Level}

At the software level, the execution proceeds through several distinct computational phases, each with different performance characteristics.

\subsubsection{Startup Phase (Configuration and Offline Computation)}

When the executable launches, the first operation is parsing the YAML configuration file.  The yaml-cpp library reads the file into a tree structure, then each value is extracted and stored in the \texttt{Config} struct.  This involves heap allocation for strings (rotor names) and dynamic arrays (rotor list).  On a modern x86-64 CPU, this takes approximately 1--5 ms depending on filesystem cache state.

Next, the control allocation matrix is built.  This is a sequence of vector cross products and scalar multiplications --- pure arithmetic on 3-element vectors.  The entire computation fits in L1 cache and completes in microseconds.

The trim solver assembles the 4$\times$4 KKT matrix and calls Eigen's QR decomposition.  For a matrix this small, the decomposition is dominated by function call overhead rather than actual floating-point operations.  Total time: approximately 10 $\mu$s.

The INDI controller builds $\bm{G} = \bJ^{-1}\bB_\tau$ (a single $3 \times 3$ matrix multiply) and inverts it.  The $3 \times 3$ inverse requires approximately 40 floating-point operations.  The condition number is computed via SVD (approximately 100 operations for a $3 \times 3$ matrix).  Total: under 1 $\mu$s.  Compare with the old CARE solver, which required 15--25 iterations of $12 \times 12$ matrix inversion.  The Butterworth filter coefficients are computed from $\gamma = \tan(\pi f_c / f_s)$ --- a single transcendental function evaluation plus a few multiplications.

\subsubsection{Runtime Phase (Simulation Loop)}

The simulation loop runs for 10{,}001 timesteps.  Each timestep involves:

\paragraph{State extraction (under 100 ns):}  Reading 13 doubles from the state vector, computing atan2 and asin for Euler angle extraction.  These transcendental functions are the most expensive single operations, taking approximately 20 ns each on modern hardware.

\paragraph{INDI control computation (under 500 ns):}  Three stages: (a) Butterworth filter update: 3 axes $\times$ 5 multiply-adds; (b) attitude and rate P loops: 6 multiplications each; (c) $\bm{G}^{-1}$ multiply: $3 \times 3$ matrix-vector (9 multiply-adds).  The priority allocator adds conditional logic (branching) for the RP/yaw decomposition, which may cause a branch misprediction on the first saturated timestep but is otherwise predictable.

\paragraph{Motor model (under 50 ns):}  3 multiply-adds (one per motor) plus 3 clamp operations.

\paragraph{PID computation (under 50 ns):}  5 multiplications, 3 additions, 2 clamp operations.  Plus the tilt compensation: 2 cosine evaluations ($\sim$40 ns) and a division.

\paragraph{RK4 integration (approximately 5 $\mu$s):}  Four evaluations of the dynamics function, each involving: one quaternion-to-rotation-matrix conversion (9 multiplications, 6 additions), one 3$\times$3 matrix-vector multiply (inertia tensor), one cross product, and one 3$\times$3 matrix inverse (precomputed).  The RK4 weighted sum is 13 multiply-adds per stage.  Total per timestep: approximately 400 floating-point operations.

\paragraph{CSV logging (approximately 10 $\mu$s):}  String formatting with \texttt{std::fixed} and \texttt{std::setprecision} is surprisingly expensive compared to the numeric computation.  File I/O is buffered by the OS (typically 4 KB or 8 KB buffer), so actual disk writes occur only when the buffer fills.

\paragraph{Total per timestep:}  Approximately 15--20 $\mu$s, dominated by CSV logging.  Without logging, the pure computation takes approximately 6--7 $\mu$s.

\paragraph{Memory usage:}  The state vector (13 doubles = 104 bytes), control vector (3 doubles = 24 bytes), motor model state (3 doubles = 24 bytes), filter state (4 doubles $\times$ 3 axes = 96 bytes), and all matrices fit in L1 cache.  The RK4 intermediates (k1--k4, each 13 doubles) add another 416 bytes.  Total working set: under 2.5 KB.  There are zero heap allocations in the hot loop.

\subsection{What Happens at the Hardware Level}

\subsubsection{CPU Execution}

On a modern out-of-order CPU (e.g., AMD Zen 4, Intel Raptor Lake), the simulation loop exhibits near-perfect instruction-level parallelism.  The Eigen matrix operations compile to packed SSE2 or AVX2 instructions: a single \texttt{vfmadd231pd} instruction performs 4 fused multiply-add operations simultaneously on 4 doubles packed in a 256-bit YMM register.

The $3\times3$ $\bm{G}^{-1}$ multiply compiles to approximately 3 \texttt{vmulpd} and 2 \texttt{vaddpd} instructions, completing in approximately 2 clock cycles on a superscalar pipeline.  The Butterworth filter update is a sequence of 5 multiply-accumulate operations per axis, also fully pipelined.  Branch prediction is essentially perfect because the simulation loop has no data-dependent branches (the only conditional is the priority allocator's saturation check, which is predictable after the first few timesteps).

\subsubsection{Memory Hierarchy}

The entire computation fits in L1 data cache (32--48 KB).  The state vector, control vector, gain matrix, inertia tensor, and all intermediates total under 2 KB.  Cache miss rate is effectively zero during the simulation loop.  The only cache misses occur during CSV output, when the output buffer is flushed to L2/L3 and eventually to main memory.

\subsubsection{Floating-Point Considerations}

All computations use IEEE 754 double-precision (64-bit) floating point, providing approximately 15--16 significant decimal digits.  The most numerically sensitive operation is the $3 \times 3$ matrix inverse of $\bm{G}$.  With a condition number of 49.25, we lose approximately $\log_{10}(49.25) \approx 1.7$ digits of precision in the inverse, leaving approximately 14 significant digits --- more than sufficient.  Compare with the old CARE solver, which required inverting a $12 \times 12$ Hamiltonian with much higher condition numbers.

The quaternion normalisation prevents the gradual accumulation of unit-norm constraint violation.  Without normalisation, the quaternion norm drifts at a rate of approximately $10^{-12}$ per timestep, accumulating to approximately $10^{-8}$ over 10{,}000 steps.  While this is small, it compounds nonlinearly: a non-unit quaternion produces a non-orthogonal rotation matrix, which violates the rigid body assumption and causes energy drift.

\subsubsection{Disk I/O}

The simulation produces a CSV file of approximately 2--3 MB (10{,}001 rows $\times$ 17 columns $\times$ approximately 15 characters per value).  The OS filesystem layer buffers writes in page cache (4 KB pages), flushing to the storage device asynchronously.  On an SSD, the final flush takes under 1 ms.  On a spinning HDD, it could take 10--50 ms due to seek latency, but this occurs only once at the end of the simulation (when the \texttt{ofstream} destructor calls \texttt{close()}).

\subsubsection{Compilation and Optimisation}

The CMake build uses \texttt{-O2} optimisation, which enables:
\begin{itemize}
    \item \textbf{Function inlining:}  Small Eigen operations (vector addition, scalar multiplication) are inlined into the caller, eliminating function call overhead.
    \item \textbf{Loop unrolling:}  Fixed-size Eigen loops (e.g., the 3-element inner products) are fully unrolled at compile time.
    \item \textbf{SIMD vectorisation:}  GCC/Clang auto-vectorise the Eigen operations when the data is aligned (Eigen's fixed-size types are 16-byte aligned by default).
    \item \textbf{Dead code elimination:}  The unused \texttt{(void)pos;} in the dynamics function is compiled to nothing.
\end{itemize}

The total compilation time (including FetchContent downloads of Eigen and yaml-cpp) is approximately 30--90 seconds on a modern machine, dominated by compiling the Eigen headers (which are heavily templated) and yaml-cpp's C++ sources.

% =============================================================================
\section{Conclusion}
% =============================================================================

We have built a complete INDI-based flight controller for a servoless asymmetric tricopter, evolving through three architectural generations: a 6-state LQR (unstable for this vehicle), a 4-state LQR with yaw damper (stable but model-dependent), and finally INDI (robust, sensor-based, and well-suited to the vehicle's capabilities).

\subsection{The INDI Architecture}

The final controller consists of:

\begin{enumerate}
    \item \textbf{INDI Controller:}  A 3-loop cascaded structure (attitude P $\to$ rate P $\to$ incremental inversion) that requires only the $3 \times 3$ control effectiveness matrix $\bm{G} = \bJ^{-1}\bB_\tau$.  No CARE equation, no Hamiltonian, no Riccati solver.  The only offline computation is a $3 \times 3$ matrix inverse.

    \item \textbf{2nd-Order Butterworth Filter:}  Estimates angular acceleration from gyroscope data at 30 Hz cutoff, providing the feedback signal for INDI's incremental law.  Pre-seeded with roll/pitch trim values (yaw deliberately excluded to prevent thrust collapse).

    \item \textbf{First-Order Motor Model:}  Simulates the 30 ms motor lag, tracking $\bu_{\text{actual}}$ for correct INDI incremental computation.

    \item \textbf{Priority-Based Allocator:}  Guarantees roll/pitch authority by sacrificing yaw when motors saturate.  Yaw was scaled in 99.95\% of timesteps --- confirming the vehicle's fundamental yaw weakness.

    \item \textbf{Altitude PID with Post-Allocation Collective:}  Maintains altitude independently of attitude control, with tilt compensation.

    \item \textbf{Heading-Aware Command:}  Ready for future position control.
\end{enumerate}

\subsection{Key Results}

\begin{itemize}
    \item \textbf{Stable hover:} Roll 5° $\to$ 1.28°, pitch 3° $\to$ 0.35° over 10 seconds.
    \item \textbf{Altitude hold:} $z \approx 0$ throughout, thrust stable at $\approx mg = 14.715$ N.
    \item \textbf{Yaw rate bounded:} Yaw rate reaches $\sim$24 rad/s by $t = 10$ s (inherent to servoless tricopter physics).
    \item \textbf{G matrix condition number:} 49.25 (yaw $\sim$49$\times$ weaker than roll/pitch).
    \item \textbf{Butterworth filter:} 30 Hz cutoff, $b_0 = 0.00782$, settling in $\sim$50 ms.
    \item \textbf{Motor model:} $\tau = 30$ ms, correctly tracked via $\bu_{\text{prev}} = \bu_{\text{actual}}$.
    \item \textbf{Priority allocator:} Yaw scaled 99.95\% of timesteps, $\alpha_{\min} = 0.003$.
    \item \textbf{All 16 unit tests passing.}
\end{itemize}

\subsection{The Critical Bug and Its Fix}

The most important engineering lesson was the thrust collapse bug (Section~\ref{sec:thrust_collapse}).  Pre-seeding the Butterworth filter with the full trim angular acceleration (including a 2.275 rad/s$^2$ yaw component) caused INDI to demand enormous motor changes through $\bm{G}^{-1}$'s amplified yaw column.  The fix was twofold: (1) pre-seed only RP, let yaw warm up naturally; (2) apply collective thrust correction \emph{after} attitude allocation, not during.  This separation of concerns --- attitude allocation is independent of thrust control --- is a fundamental principle for robust multirotor flight control.

\subsection{The Physical Lesson}

The deepest lesson from this project is that good control design starts with understanding the physics.  A mathematically optimal controller (LQR) can be physically inappropriate if it asks the vehicle to do something the hardware cannot achieve.  INDI is better matched to this vehicle because it works \emph{with} the measurements rather than \emph{against} the model.  It measures what is happening, computes what needs to change, and lets the physics take care of the rest.  For a servoless tricopter with its inherent yaw weakness and residual torques, this incremental, measurement-based approach is far more robust than any model-based optimal controller.

\end{document}
